---
title: "Introduction to ADNIMetabo"
author: 
  - name: Dany Mukesha
    affiliation:
    - Université Côte d'Azur
    email: danymukesha@gmail.com
abstract:
  We provide a collection of datasets from the Alzheimer's Disease 
  Neuroimaging Initiative (ADNI), a landmark study aimed at understanding 
  the progression of Alzheimer's disease. This package includes various 
  types of data, such as demographic information, clinical assessments, 
  biomarker measurements, imaging data, and cognitive composite scores. 
  The datasets are carefully curated and documented to facilitate easy 
  access and analysis for researchers and clinicians. This vignette guides 
  users through the process of loading and utilizing these datasets, 
  providing examples of how to explore and analyze the data.
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('ADNIMetabo')`"
vignette: >
  %\VignetteIndexEntry{Introduction to ADNIMetabo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL
)
```
# Introduction

Alzheimer's disease is a complex and multifactorial neurodegenerative 
disorder that affects millions of people worldwide. The Alzheimer's Disease 
Neuroimaging Initiative (ADNI) is a multisite study launched in 2003 
to collect and share data that can help researchers understand the progression 
of Alzheimer's disease. The `ADNIMetabo` package is developed to make the ADNI 
datasets readily available and easily accessible for analysis.

In this package, it includes a range of datasets, each focusing on different 
aspects of the ADNI study:

- **Demographic data**: Participant demographic information such as age, sex, 
education, ethnicity, and marital status.
- **Clinical assessments**: Baseline and follow-up clinical data, including 
diagnosis, CDR-SB, ADAS scores, MMSE, and RAVLT scores.
- **Biomarker data**: Measurements of CSF ABETA, TAU, PTAU, FDG-PET, PIB, 
and AV45.
- **Imaging data**: MRI field strength, FreeSurfer software version, and 
various brain region volumes.
- **Cognitive domposite scores**: ADNI modified Preclinical Alzheimer's 
Cognitive Composite (PACC) scores with Digit Symbol Substitution and Trails B.
- **ECog scores**: ECog scores for both participants and study partners.

This document will walk you through the steps to install and load the
`ADNIMetabo` package, access the different datasets, and provide examples 
of how to explore and analyze the data. It is intended to serve as a guide 
for researchers, clinicians, and students who are interested in using the 
ADNI datasets for their studies.

We hope that by using the `ADNIMetabo` package, users can take advantage of
the extensive and well-curated datasets from the ADNI study to advance our
understanding of Alzheimer's disease and contribute to the development of 
diagnostic and therapeutic strategies.

Let's start with get access to various datasets from ADNI:

```{r installation, eval=FALSE, include=TRUE}
devtools::install_github("danymukesha/ADNIData") #for installation
```

```{r load_data}
library(ADNIMetabo)

data("demographic_data", package = "ADNIMetabo")
data("baseline_clinical_data", package = "ADNIMetabo")
data("follow_up_clinical_data", package = "ADNIMetabo")
data("biomarker_data", package = "ADNIMetabo")
data("imaging_data", package = "ADNIMetabo")
data("cognitive_composite_data", package = "ADNIMetabo")
data("ecog_scores_data", package = "ADNIMetabo")
```

## Exploring the datasets

an example of how you can explore the demographic data:

```{r view_demographic_data}
head(demographic_data)
```

## Short analysis

### Demographic summary

```{r demographic_summary}
# summarize age and education
summary(demographic_data$AGE)
summary(demographic_data$PTEDUCAT)

# count the number of participants by sex and ethnicity
table(demographic_data$PTGENDER)
table(demographic_data$PTETHCAT)
```

### Clinical assessments


```{r clinical_assessments}
# summarize baseline CDR-SB and MMSE scores
summary(baseline_clinical_data$CDRSB_bl)
summary(baseline_clinical_data$MMSE_bl)

# plot the distribution of baseline ADAS 11 scores
hist(baseline_clinical_data$ADAS11_bl, main = "Baseline ADAS 11 Scores", xlab = "ADAS 11 Score")
```

### Biomarker analysis


```{r biomarker_analysis}
# summarize CSF ABETA and TAU levels
summary(biomarker_data$ABETA)
summary(biomarker_data$TAU)

# plot the relationship between CSF ABETA and TAU
plot(biomarker_data$ABETA, biomarker_data$TAU, main = "CSF ABETA vs TAU", xlab = "CSF ABETA", ylab = "CSF TAU")
```

## Short message

The `ADNIMetabo` package will provide a set of datasets from the ADNI study, 
making it easier to analyze and understand the progression of Alzheimer's 
disease. This vignette has shown you how to load and explore these datasets, a
nd how to perform basic analyses.

For more detailed documentation on each dataset, please refer to the respective
help files:

```r
?demographic_data
?baseline_clinical_data
?follow_up_clinical_data
?biomarker_data
?imaging_data
?cognitive_composite_data
?ecog_scores_data
```

## Acknowledgments

The data included in this package is from the Alzheimer's Disease Neuroimaging
Initiative (ADNI) database. ADNI is a multisite study launched in 2003 by the 
National Institute on Aging (NIA), the National Institute of Biomedical Imaging
and Bioengineering (NIBIB), the Food and Drug Administration (FDA), private 
pharmaceutical companies, and non-profit organizations.

```{r reproduce3, echo=FALSE}
library("sessioninfo")
options(width = 120)
session_info()
```
