---
title: 'ADNI Baseline Analysis: AD vs CN'
# subtitle: "Case-Control Study"
author: "Dany MUKESHA"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
    fig_width: 10
    fig_height: 6
  word_document:
    toc: true
  pdf_document:
    toc: true
    fig_width: 8
    fig_height: 5
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE, 
    warning = FALSE, 
    message = FALSE, 
    fig.align = "center", 
    dpi = 300)

library(tidyverse)
library(ggplot2)
library(ggrepel)
library(gridExtra)
library(grid)
library(corrplot)
library(tableone)
library(RColorBrewer)
library(viridis)
library(plotly)
library(DT)
library(kableExtra)
library(ggpubr)
library(broom)
library(car)
library(effsize)
devtools::load_all()

theme_set(theme_bw() + 
          theme(plot.title = element_text(hjust = 0.5, face = "bold",
                                          size = 14, vjust = 2.4),
                legend.position = "bottom",
                strip.text = element_text(face = "bold", size = 12.5),
                axis.text = element_text(size = 11, face = "plain"),
                axis.title = element_text(size = 12)))

my_comparisons <- list( c("AD", "CN"))

stat_comp <- stat_compare_means(label = "p.signif", method = "t.test",
                                vjust = 0.6, ref.group = "0.00001", 
                                comparisons = my_comparisons) 
custom_colors <- c("#2E86AB", "#A23B72", "#F18F01", "#C73E1D")
```

# Executive Summary

This analysis examines baseline differences between Alzheimer's Disease
(AD) patients and Cognitively Normal (CN) controls in the ADNI dataset.
The study includes demographic, clinical, neuropsychological, and
biomarker comparisons.

# Data Preparation

```{r get-rids-with-metabo-data, eval=TRUE, include=FALSE}
data <- data.table::fread("03_EffectsAdjusted.csv", sep = ",") 
RIDs_with_metabo_data <- data$RID
```

```{r data-prep}
# baseline subset for case-control analysis
baseline_data <- adnim %>%
  filter(VISCODE == "bl", DX_bl %in% c("AD", "CN")) %>%
  mutate(
    Group = factor(DX_bl, levels = c("CN", "AD")),
    APOE4_status = case_when(
      APOE4 == 0 ~ "No APOE4",
      APOE4 == 1 ~ "One APOE4",
      APOE4 == 2 ~ "Two APOE4",
      TRUE ~ "Unknown"
    ),
    Education_level = case_when(
      PTEDUCAT < 12 ~ "< High School",
      PTEDUCAT == 12 ~ "High School",
      PTEDUCAT > 12 & PTEDUCAT < 16 ~ "Some College",
      PTEDUCAT >= 16 ~ "College+",
      TRUE ~ "Unknown"
    ),
    ABETA_numeric = as.numeric(ifelse(ABETA_bl == "", NA, ABETA_bl)),
    TAU_numeric = as.numeric(ifelse(TAU_bl == "", NA, TAU_bl))
  ) %>%
  dplyr::filter(RID %in% RIDs_with_metabo_data) %>%
  select(-ABETA_bl, -TAU_bl, -ABETA, -TAU) 

cat("Sample Sizes:\n")
table(baseline_data$Group)
```

# Demographic Characteristics

## Table 1: Baseline Demographics and Clinical Characteristics

```{r demographics-table}
demo_vars <- c("AGE", "PTGENDER", "PTEDUCAT", "PTRACCAT", "Education_level", 
               "APOE4", "APOE4_status")

tab1 <- CreateTableOne(vars = demo_vars, 
                       strata = "Group", 
                       data = baseline_data,
                       test = TRUE)

kable(print(tab1, showAllLevels = TRUE, printToggle = FALSE), 
      caption = "Table 1: Baseline Demographics") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Demographic Visualizations

```{r demo-plots, fig.height=8, fig.width=12}
p1 <- ggplot(baseline_data, aes(x = Group, y = AGE, fill = Group)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3) +
  scale_fill_manual(values = custom_colors[1:2]) +
  labs(title = "Age Distribution by Group", y = "Age (years)") +
  theme(legend.position = "none")

p2 <- baseline_data %>%
  count(Group, PTGENDER) %>%
  group_by(Group) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = Group, y = prop, fill = PTGENDER)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = custom_colors[3:4]) +
  labs(title = "Gender Distribution by Group", y = "Proportion", fill = "Gender")

p3 <- baseline_data %>%
  count(Group, Education_level) %>%
  ggplot(aes(x = Education_level, y = n, fill = Group)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = custom_colors[1:2]) +
  labs(title = "Education Distribution by Group", 
       x = "Education Level", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p4 <- baseline_data %>%
  count(Group, APOE4_status) %>%
  group_by(Group) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = APOE4_status, y = prop, fill = Group)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = custom_colors[1:2]) +
  labs(title = "APOE4 Status Distribution by Group", 
       x = "APOE4 Status", y = "Proportion") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

grid.arrange(p1, p2, p3, p4, ncol = 2)
```

# Neuropsychological Assessment Comparisons

## Table 2: Cognitive Test Scores

```{r cognitive-table}
cog_vars <- c("MMSE_bl", "ADAS11_bl", "ADAS13_bl", "ADASQ4_bl", 
              "LDELTOTAL_BL", "RAVLT_immediate_bl", "RAVLT_learning_bl", 
              "RAVLT_forgetting_bl", "MOCA_bl", "CDRSB_bl", "DIGITSCOR_bl", "FAQ_bl")

tab2 <- CreateTableOne(vars = cog_vars, 
                       strata = "Group", 
                       data = baseline_data,
                       test = TRUE,
                       testNonNormal = TRUE)

kable(print(tab2, printToggle = FALSE), 
      caption = "Table 2: Neuropsychological Assessment Scores") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Cognitive Assessment Visualizations

```{r cognitive-plots, fig.height=10, fig.width=14}
cog_data <- baseline_data %>%
  select(Group, all_of(cog_vars)) %>%
  gather(key = "Test", value = "Score", -Group) %>%
  filter(!is.na(Score)) %>%
  mutate(
    Test_Label = case_when(
      Test == "MMSE_bl" ~ "MMSE",
      Test == "ADAS11_bl" ~ "ADAS-11",
      Test == "ADAS13_bl" ~ "ADAS-13", 
      Test == "ADASQ4_bl" ~ "ADAS-Q4",
      Test == "LDELTOTAL_BL" ~ "Logical Memory",
      Test == "RAVLT_immediate_bl" ~ "RAVLT Immediate",
      Test == "RAVLT_learning_bl" ~ "RAVLT Learning",
      Test == "RAVLT_forgetting_bl" ~ "RAVLT Forgetting",
      Test == "MOCA_bl" ~ "MOCA",
      Test == "CDRSB_bl" ~ "CDR-SB",
      Test == "DIGITSCOR_bl" ~ "Digit Symbol Substitution",
      Test == "FAQ_bl" ~ "FAQ",
      TRUE ~ Test
    )
  )

my_comparisons <- list( c("AD", "CN"))

p <- ggplot(cog_data, aes(x = Group, y = Score, fill = Group)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 0.5) +
  scale_fill_manual(values = custom_colors[1:2]) +
  facet_wrap(~ Test_Label, scales = "free_y", ncol = 4) +
  labs(title = "Cognitive Test Scores by Diagnosis", 
       y = "Score", x = "Diagnosis") +
  theme(legend.position = "none", plot.title = element_text(size = 22, vjust = 1)) +
  # stat_compare_means(comparisons = my_comparisons) +
  # stat_compare_means(label.y = 45) +
  # stat_compare_means(method = "anova", label.y = 40) +      # Add global p-value
  stat_compare_means(label = "p.signif", method = "t.test",vjust = 0.6,
                     ref.group = "0.00001", comparisons = my_comparisons) 
p
ggsave("results_baseline_AD-vs-CN/cog_tests_by_dx.png", 
       p, height = 10, width = 14, dpi = 300)
```

## Effect Sizes for Cognitive Tests

```{r effect-sizes}
# Calculate effect sizes (Cohen's d)
effect_sizes <- cog_data %>%
  group_by(Test_Label) %>%
  summarise(
    cohens_d = cohen.d(Score ~ Group, na.rm = TRUE)$estimate,
    ci_lower = cohen.d(Score ~ Group, na.rm = TRUE)$conf.int[1],
    ci_upper = cohen.d(Score ~ Group, na.rm = TRUE)$conf.int[2],
    p_value = t.test(Score ~ Group, na.rm = TRUE)$p.value,
    .groups = 'drop'
  ) %>%
  arrange(desc(abs(cohens_d)))

# Plot effect sizes
p <- ggplot(effect_sizes, aes(x = reorder(Test_Label, abs(cohens_d)), y = cohens_d)) +
  geom_col(aes(fill = abs(cohens_d) > 0.8), alpha = 0.8) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
  coord_flip() +
  scale_fill_manual(values = c("FALSE" = custom_colors[1], "TRUE" = custom_colors[2]),
                   name = "Large Effect\n(|d| > 0.8)") +
  labs(title = "Effect Sizes (Cohen's d) for Cognitive Tests", 
       subtitle = "AD vs CN Comparison",
       x = "Cognitive Test", y = "Cohen's d (95% CI)") +
  geom_hline(yintercept = c(-0.8, -0.5, -0.2, 0.2, 0.5, 0.8), 
             linetype = "dashed", alpha = 0.5)

p
ggsave("results_baseline_AD-vs-CN/cog_tests_effect_sizes.png", 
       p, height = 10, width = 14, dpi = 300)

effect_sizes_formatted <- effect_sizes %>%
    arrange(desc(abs(cohens_d))) %>%
    mutate(CI = sprintf("[%.2f, %.2f]", ci_lower, ci_upper),
           d = sprintf("%.2f", cohens_d)) %>%
    select(Test = Test_Label, `Cohen's d` = d, `95% CI` = CI)

kable(effect_sizes_formatted, digits = 3, 
      caption = "Effect Sizes for Cognitive Test Differences") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# Biomarker Analysis

## Table 3: Biomarker Comparisons

```{r biomarker-table}
bio_vars <- c("FDG_bl", "AV45_bl", "ABETA_numeric", "TAU_numeric", "ICV_bl")

tab3 <- CreateTableOne(vars = bio_vars, 
                       strata = "Group", 
                       data = baseline_data,
                       test = TRUE)

kable(print(tab3, printToggle = FALSE), 
      caption = "Table 3: Biomarker Comparisons") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Biomarker Visualizations

```{r biomarker-plots, fig.height=8, fig.width=12}
bio_data <- baseline_data %>%
  select(Group, FDG_bl, AV45_bl, ABETA_numeric, TAU_numeric) %>%
  gather(key = "Biomarker", value = "Value", -Group) %>%
  filter(!is.na(Value)) %>%
  mutate(
    Biomarker_Label = case_when(
      Biomarker == "FDG_bl" ~ "FDG-PET",
      Biomarker == "AV45_bl" ~ "Amyloid PET",
      Biomarker == "ABETA_numeric" ~ "CSF Aβ42",
      Biomarker == "TAU_numeric" ~ "CSF Tau",
      TRUE ~ Biomarker
    )
  )

p1 <- ggplot(bio_data, aes(x = Group, y = Value, fill = Group)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 2) +
  scale_fill_manual(values = custom_colors[1:2]) +
  facet_wrap(~ Biomarker_Label, scales = "free_y", ncol = 2) +
  labs(title = "Biomarker Levels by Diagnosis", 
       y = "Value", x = "Diagnosis") +
  theme(legend.position = "none", plot.title = element_text(size = 22, vjust = 1)) +
  stat_comp

print(p1)
ggsave("results_baseline_AD-vs-CN/csf_biomarkers_by_dx.png", 
       p1, height = 8, width = 12, dpi = 300)
```


```{r biomarker-plots_corr, fig.height=10, fig.width=10}
bio_corr_data <- baseline_data %>%
  select(FDG_bl, AV45_bl, ABETA_numeric, TAU_numeric) %>%
  rename("FDG-PET" = FDG_bl, 
         "Amyloid PET" = AV45_bl,
         "CSF A\u03B242" = ABETA_numeric,
         "CSF Tau" = TAU_numeric)

# png("results_baseline_AD-vs-CN/biomarker_corrplot.png", width = 1200, height = 1000, res = 150)
corrplot(
  cor(bio_corr_data, use = "complete.obs"), method = "color",
  order = "hclust", addCoef.col = "black", tl.col = "black", tl.srt = 45,           
  number.cex = 1.4, tl.cex = 1.2, sig.level = 0.01, cex.main = 1.8,       
  insig = "blank", diag = FALSE, title = "Biomarker Correlations",
  mar = c(1, 1, 3, 1)    # margin: bottom, left, top, right
)
# dev.off()
```

## Effect Sizes for Biomarkers

```{r effect-size_bios}
# Calculate effect sizes (Cohen's d)
bio_effect_sizes <- bio_data %>%
  group_by(Biomarker_Label) %>%
  summarise(
    cohens_d = cohen.d(Value ~ Group, na.rm = TRUE)$estimate,
    ci_lower = cohen.d(Value ~ Group, na.rm = TRUE)$conf.int[1],
    ci_upper = cohen.d(Value ~ Group, na.rm = TRUE)$conf.int[2],
    p_value = t.test(Value ~ Group, na.rm = TRUE)$p.value,
    .groups = 'drop'
  ) %>%
  arrange(desc(abs(cohens_d)))

# Plot effect sizes
p <- ggplot(bio_effect_sizes, aes(x = reorder(Biomarker_Label, abs(cohens_d)), y = cohens_d)) +
  geom_col(aes(fill = abs(cohens_d) > 0.8), alpha = 0.8) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
  coord_flip() +
  scale_fill_manual(values = c("FALSE" = custom_colors[1], "TRUE" = custom_colors[2]),
                   name = "Large Effect\n(|d| > 0.8)") +
  labs(title = "Effect Sizes (Cohen's d) for Biomarkers Values", 
       subtitle = "AD vs CN Comparison",
       x = "Cognitive Test", y = "Cohen's d (95% CI)") +
  geom_hline(yintercept = c(-0.8, -0.5, -0.2, 0.2, 0.5, 0.8), 
             linetype = "dashed", alpha = 0.5)

p
ggsave("results_baseline_AD-vs-CN/biomarkers_effect_sizes.png", 
       p, height = 10, width = 14, dpi = 300)

bio_effect_sizes_formatted <- bio_effect_sizes %>%
    arrange(desc(abs(cohens_d))) %>%
    mutate(CI = sprintf("[%.2f, %.2f]", ci_lower, ci_upper),
           d = sprintf("%.2f", cohens_d)) %>%
    select(Test = Biomarker_Label, `Cohen's d` = d, `95% CI` = CI)

kable(bio_effect_sizes_formatted, digits = 3, 
      caption = "Effect Sizes for Biomarkers Level Differences") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# Neuroimaging Analysis: Structural MRI

## Overview of Neuroimaging Data

The ADNI dataset also includes structural MRI measurements derived from
FreeSurfer analysis, providing detailed information about brain regional
volumes and cortical thickness measures.

```{r imaging-data-prep}
# add imaging variables to baseline data
baseline_data <- baseline_data %>%
  mutate(
    # first: normalize brain volumes by ICV to account for head size differences
    # second: create atrophy indices (higher values = more atrophy)
    # thirf: we perform entricle-to-brain ratio (VBR)
    Ventricles_norm = Ventricles_bl / ICV_bl * 1000,
    Hippocampus_norm = Hippocampus_bl / ICV_bl * 1000,
    WholeBrain_norm = WholeBrain_bl / ICV_bl * 1000,
    Entorhinal_norm = Entorhinal_bl / ICV_bl * 1000,
    Fusiform_norm = Fusiform_bl / ICV_bl * 1000,
    MidTemp_norm = MidTemp_bl / ICV_bl * 1000,
    
    Hippocampal_Atrophy = 1 - (Hippocampus_norm / max(Hippocampus_norm, na.rm = TRUE)),
    Entorhinal_Atrophy = 1 - (Entorhinal_norm / max(Entorhinal_norm, na.rm = TRUE)),
    
    VBR = Ventricles_bl / WholeBrain_bl
  )

imaging_summary <- baseline_data %>%
  summarise(
    across(c(Ventricles_bl, Hippocampus_bl, WholeBrain_bl, 
             Entorhinal_bl, Fusiform_bl, MidTemp_bl, ICV_bl), 
           ~sum(!is.na(.)), .names = "n_{.col}"),
    .by = Group
  )

kable(imaging_summary, caption = "Structural MRI Data Availability") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Table 4: Structural MRI Comparisons

```{r imaging-table}
imaging_vars <- c("ICV_bl", "Ventricles_norm", "Hippocampus_norm", "WholeBrain_norm", 
                  "Entorhinal_norm", "Fusiform_norm", "MidTemp_norm", "VBR")

tab4 <- CreateTableOne(vars = imaging_vars, 
                       strata = "Group", 
                       data = baseline_data,
                       test = TRUE)

kable(print(tab4, printToggle = FALSE), 
      caption = "Table 4: Structural MRI Measurements (ICV-normalized)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Regional Brain Volume Comparisons

```{r regional-volumes, fig.height=10, fig.width=14}
regional_data <- baseline_data %>%
  select(Group, Ventricles_norm, Hippocampus_norm, WholeBrain_norm, 
         Entorhinal_norm, Fusiform_norm, MidTemp_norm) %>%
  gather(key = "Region", value = "Volume", -Group) %>%
  filter(!is.na(Volume)) %>%
  mutate(
    Region_Label = case_when(
      Region == "Ventricles_norm" ~ "Ventricles",
      Region == "Hippocampus_norm" ~ "Hippocampus",
      Region == "WholeBrain_norm" ~ "Whole Brain",
      Region == "Entorhinal_norm" ~ "Entorhinal Cortex",
      Region == "Fusiform_norm" ~ "Fusiform Gyrus",
      Region == "MidTemp_norm" ~ "Middle Temporal",
      TRUE ~ Region
    ),
    Region_Category = case_when(
      Region %in% c("Hippocampus_norm", "Entorhinal_norm") ~ "Memory-Related",
      Region %in% c("Fusiform_norm", "MidTemp_norm") ~ "Temporal Cortex",
      Region == "WholeBrain_norm" ~ "Global Measure",
      Region == "Ventricles_norm" ~ "CSF Space",
      TRUE ~ "Other"
    )
  )

p1 <- ggplot(regional_data, aes(x = Group, y = Volume, fill = Group)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  geom_jitter(width = 0.2, alpha = 0.2, size = 1.5) +
  scale_fill_manual(values = custom_colors[1:2]) +
  facet_wrap(~ Region_Label, scales = "free_y", ncol = 3) +
  labs(title = "Regional Brain Volumes by Diagnostic Group", 
       subtitle = "Volumes normalized by intracranial volume (ICV)",
       y = "Normalized Volume (per 1000 ICV units)") +
  theme(legend.position = "none", 
        strip.text = element_text(face = "bold", size = 15),
        axis.text = element_text(size = 14, face = "plain"),
        plot.title = element_text(size = 22, vjust = 1),
        axis.title = element_text(size = 14)) +
  stat_compare_means(label = "p.signif", method = "t.test",
                     ref.group = "0.00001", comparisons = my_comparisons) 

print(p1)
ggsave("results_baseline_AD-vs-CN/brain_region_by_dx.png", 
       p1, height = 8, width = 12, dpi = 300)

# effect sizes for brain regions
regional_effects <- regional_data %>%
  group_by(Region_Label, Region_Category) %>%
  summarise(
    cohens_d = cohen.d(Volume ~ Group, na.rm = TRUE)$estimate,
    ci_lower = cohen.d(Volume ~ Group, na.rm = TRUE)$conf.int[1],
    ci_upper = cohen.d(Volume ~ Group, na.rm = TRUE)$conf.int[2],
    p_value = t.test(Volume ~ Group, na.rm = TRUE)$p.value,
    .groups = 'drop'
  ) %>%
  arrange(desc(abs(cohens_d)))

p2 <- ggplot(regional_effects, aes(x = reorder(Region_Label, abs(cohens_d)),
                                   y = cohens_d)) +
  geom_col(aes(fill = Region_Category), alpha = 0.8) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
  coord_flip() +
  scale_fill_viridis_d(name = "Region Category") +
  labs(title = "Effect Sizes for Regional Brain Volume Differences", 
       subtitle = "AD vs CN Comparison (Cohen's d with 95% CI)",
       x = "Brain Region", y = "Cohen's d") +
  geom_hline(yintercept = c(-1.2, -0.8, -0.5, -0.2, 0, 0.2, 0.5, 0.8, 1.2), 
             linetype = "dashed", alpha = 0.3) +
  theme(strip.text = element_text(face = "bold", size = 15),
        axis.text = element_text(size = 14, face = "plain"),
        plot.title = element_text(size = 22, vjust = 1),
        axis.title = element_text(size = 13), 
        legend.text = element_text(size = 10))

print(p2)
ggsave("results_baseline_AD-vs-CN/brain_region_effect_sizes.png", 
       p2, height = 6, width = 12, dpi = 300)

regional_effects_formatted <- regional_effects %>%
    arrange(desc(abs(cohens_d))) %>%
    mutate(CI = sprintf("[%.2f, %.2f]", ci_lower, ci_upper),
           d = sprintf("%.2f", cohens_d)) %>%
    select(Test = Region_Label, `Cohen's d` = d, `95% CI` = CI)

kable(regional_effects_formatted, digits = 3, 
      caption = "Effect Sizes for Regional Brain Volume Differences") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Advanced Neuroimaging Analysis

### Brain Volume Ratios and Composite Scores

```{r brain-ratios, fig.height=8, fig.width=12}
# we meaningful brain ratios
baseline_data <- baseline_data %>%
  mutate(
    # first: hippocampus to total brain ratio
    # second: we do entorhinal to hippocampus ratio
    # third: temporal lobe composite (fusiform + middle temporal + entorhinal + hippocampus)
    # forth: we calculate atrophy severity index combining key AD regions
    Hip_to_Brain = Hippocampus_bl / WholeBrain_bl,
    
    Ento_to_Hip = Entorhinal_bl / Hippocampus_bl,
    
    Temporal_Composite = (Fusiform_norm + MidTemp_norm + Entorhinal_norm + Hippocampus_norm),
    
    AD_Atrophy_Index = scale(Hippocampal_Atrophy)[,1] + 
                      scale(Entorhinal_Atrophy)[,1] + 
                      scale(VBR)[,1]
  )

ratio_data <- baseline_data %>%
  select(Group, Hip_to_Brain, Ento_to_Hip, VBR, AD_Atrophy_Index) %>%
  gather(key = "Ratio", value = "Value", -Group) %>%
  filter(!is.na(Value)) %>%
  mutate(
    Ratio_Label = case_when(
      Ratio == "Hip_to_Brain" ~ "Hippocampus/Brain Ratio",
      Ratio == "Ento_to_Hip" ~ "Entorhinal/Hippocampus Ratio", 
      Ratio == "VBR" ~ "Ventricle/Brain Ratio",
      Ratio == "AD_Atrophy_Index" ~ "AD Atrophy Index",
      TRUE ~ Ratio
    )
  )

p <- ggplot(ratio_data, aes(x = Group, y = Value, fill = Group)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 0.5) +
  scale_fill_manual(values = custom_colors[1:2]) +
  facet_wrap(~ Ratio_Label, scales = "free_y", ncol = 2) +
  labs(title = "Brain Volume Ratios and Composite Measures", 
       y = "Ratio/Index Value") +
  theme(legend.position = "none", 
        strip.text = element_text(face = "bold", size = 12.5),
        axis.text = element_text(size = 11, face = "plain"),
        plot.title = element_text(size = 13, vjust = 2.4),
        axis.title = element_text(size = 12)) +
  stat_compare_means(label = "p.signif", method = "t.test",
                     ref.group = "0.00001", comparisons = my_comparisons) 

ggsave("results_baseline_AD-vs-CN/brain_region_ratios.png", 
       p, height = 7, width = 10, dpi = 300)

# statistical tests for ratios
ratio_tests <- ratio_data %>%
  group_by(Ratio_Label) %>%
  summarise(
    AD_mean = mean(Value[Group == "AD"], na.rm = TRUE),
    CN_mean = mean(Value[Group == "CN"], na.rm = TRUE),
    p_value = t.test(Value ~ Group, na.rm = TRUE)$p.value,
    cohens_d = cohen.d(Value ~ Group, na.rm = TRUE)$estimate,
    ci_lower = cohen.d(Value ~ Group, na.rm = TRUE)$conf.int[1],
    ci_upper = cohen.d(Value ~ Group, na.rm = TRUE)$conf.int[2],
    .groups = 'drop'
  )

kable(ratio_tests, digits = 4, 
      caption = "Statistical Comparison of Brain Volume Ratios") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Correlation Analysis: Brain Structure and Cognition

```{r brain-cognition-correlation, fig.height=10, fig.width=14}
# we do correlation between brain measures and cognitive scores
brain_cog_data <- baseline_data %>%
  select(Group, MMSE_bl, ADAS11_bl, ADAS13_bl, LDELTOTAL_BL, RAVLT_immediate_bl,
         Hippocampus_norm, Entorhinal_norm, Fusiform_norm, MidTemp_norm, 
         VBR, AD_Atrophy_Index) %>%
  na.omit()

# corr. matrix
cor_matrix <- cor(brain_cog_data %>% select(-Group), use = "complete.obs")

corrplot(cor_matrix, method = "color", type = "upper", 
         order = "hclust", tl.cex = 0.8, tl.col = "black",
         title = "Correlations: Brain Structure and Cognitive Measures",
         mar = c(0,0,2,0))

# we show scatter plots for key relationships
key_relationships <- data.frame(
  MMSE = brain_cog_data$MMSE_bl,
  ADAS11 = brain_cog_data$ADAS11_bl,
  Hippocampus = brain_cog_data$Hippocampus_norm,
  Entorhinal = brain_cog_data$Entorhinal_norm,
  VBR = brain_cog_data$VBR,
  Group = brain_cog_data$Group
)

# MMSE vs Hippocampus
p1 <- ggplot(key_relationships, aes(x = Hippocampus, y = MMSE, color = Group)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_color_manual(values = custom_colors[1:2]) +
  labs(title = "MMSE Score vs Hippocampal Volume",
       x = "Hippocampal Volume (ICV-normalized)",
       y = "MMSE Score") +
  theme(legend.position = "bottom")

# ADAS11 vs Entorhinal
p2 <- ggplot(key_relationships, aes(x = Entorhinal, y = ADAS11, color = Group)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_color_manual(values = custom_colors[1:2]) +
  labs(title = "ADAS-11 Score vs Entorhinal Volume",
       x = "Entorhinal Volume (ICV-normalized)", 
       y = "ADAS-11 Score") +
  theme(legend.position = "bottom")

# MMSE vs VBR
p3 <- ggplot(key_relationships, aes(x = VBR, y = MMSE, color = Group)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_color_manual(values = custom_colors[1:2]) +
  labs(title = "MMSE Score vs Ventricle-to-Brain Ratio",
       x = "Ventricle-to-Brain Ratio",
       y = "MMSE Score") +
  theme(legend.position = "bottom")

# Combined atrophy index vs cognition
p4 <- ggplot(brain_cog_data, aes(x = AD_Atrophy_Index, y = MMSE_bl, color = Group)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_color_manual(values = custom_colors[1:2]) +
  labs(title = "MMSE Score vs AD Atrophy Index",
       x = "AD Atrophy Index (standardized)",
       y = "MMSE Score") +
  theme(legend.position = "bottom")

get_legend <- function(my_plot) {
  tmp <- ggplotGrob(my_plot)
  legend <- gtable::gtable_filter(tmp, "guide-box")
  return(legend)
}
legend <- get_legend(p1 + theme(legend.position = "bottom"))
p1_nolegend <- p1 + theme(legend.position = "none")
p2_nolegend <- p2 + theme(legend.position = "none")
p3_nolegend <- p3 + theme(legend.position = "none")
p4_nolegend <- p4 + theme(legend.position = "none")

combined <- grid.arrange(x,
  arrangeGrob(p1_nolegend, p2_nolegend, p3_nolegend, p4_nolegend, ncol = 2),
  legend,
  nrow = 2,
  heights = c(12, 1)  
)

ggsave("results_baseline_AD-vs-CN/brain_region_vs_cognitive.png", 
       combined, height = 9, width = 12, dpi = 300)
```

### Machine Learning Classification using Neuroimaging

```{r ml-imaging, fig.height=8, fig.width=12}
ml_data <- baseline_data %>%
  select(Group, AGE, Male, PTEDUCAT, APOE4_1, APOE4_2,
         Hippocampus_norm, Entorhinal_norm, Fusiform_norm, MidTemp_norm,
         Ventricles_norm, VBR, AD_Atrophy_Index, 
         bio_vars, CDRSB) %>%
  mutate(AD = ifelse(Group == "AD", 1, 0)) %>%
  na.omit()

# data from imaging-only model
imaging_model <- glm(AD ~ Hippocampus_norm + Entorhinal_norm + Fusiform_norm + 
                     MidTemp_norm + Ventricles_norm + VBR,
                     data = ml_data, family = "binomial")

# combined model (demographics + imaging)
combined_model <- glm(AD ~ AGE + Male + PTEDUCAT + APOE4_1 + APOE4_2 +
                      Hippocampus_norm + Entorhinal_norm + Fusiform_norm + 
                      MidTemp_norm + Ventricles_norm + VBR,
                      data = ml_data, family = "binomial")

# CSF + PET model
bio_model <- glm(AD ~ FDG_bl + AV45_bl + ABETA_numeric + TAU_numeric + ICV_bl,
                      data = ml_data, family = "binomial")

# CDR model
cdr_model <- glm(AD ~ CDRSB,
                      data = ml_data, family = "binomial")

# comparing models
imaging_results <- tidy(imaging_model) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    OR = exp(estimate),
    CI_lower = exp(estimate - 1.96 * std.error),
    CI_upper = exp(estimate + 1.96 * std.error),
    Model = "Imaging Only"
  )

combined_results <- tidy(combined_model) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    OR = exp(estimate),
    CI_lower = exp(estimate - 1.96 * std.error),
    CI_upper = exp(estimate + 1.96 * std.error),
    Model = "Combined"
  )

imaging_auc <- pROC::auc(pROC::roc(ml_data$AD, predict(imaging_model, type = "response")))
combined_auc <- pROC::auc(pROC::roc(ml_data$AD, predict(combined_model, type = "response")))
bio_auc <- pROC::auc(pROC::roc(ml_data$AD, predict(bio_model, type = "response")))
cdr_auc <- pROC::auc(pROC::roc(ml_data$AD, predict(cdr_model, type = "response")))

cat("Model Performance (AUC):\n")
cat("Imaging Only:", round(imaging_auc, 3), "\n")
cat("Combined Model:", round(combined_auc, 3), "\n")
cat("CSF + PET Model:", round(bio_auc, 3), "\n")
cat("CDR-SB Model:", round(cdr_auc, 3), "\n")

roc_imaging <- pROC::roc(ml_data$AD, predict(imaging_model, type = "response"))
roc_combined <- pROC::roc(ml_data$AD, predict(combined_model, type = "response"))
roc_cdr <- pROC::roc(ml_data$AD, predict(cdr_model, type = "response"))
roc_bio <- pROC::roc(ml_data$AD, predict(bio_model, type = "response"))


# roc_data <- data.frame(
#   Sensitivity = c(roc_imaging$sensitivities, roc_combined$sensitivities),
#   Specificity = c(1 - roc_imaging$specificities, 1 - roc_combined$specificities),
#   Model = c(rep("Imaging Only", length(roc_imaging$sensitivities)),
#             rep("Combined", length(roc_combined$sensitivities)))
# )

roc_df_imaging <- data.frame(Sensitivity = roc_imaging$sensitivities, 
  Specificity = 1 - roc_imaging$specificities, 
  Model = paste0("MRI Imaging (AUC = ", round(imaging_auc, 3), ")"))

roc_df_combined <- data.frame(Sensitivity = roc_combined$sensitivities,
  Specificity = 1 - roc_combined$specificities, 
  Model = paste0("Demographics + MRI (AUC = ", round(combined_auc, 3), ")"))

roc_df_cdr <- data.frame(Sensitivity = roc_cdr$sensitivities,
  Specificity = 1 - roc_cdr$specificities, 
  Model = paste0("CDR-SB (AUC = ", round(cdr_auc, 3), ")"))

roc_df_bio <- data.frame(Sensitivity = roc_bio$sensitivities,
  Specificity = 1 - roc_bio$specificities, 
  Model = paste0("CSF + PET (AUC = ", round(bio_auc, 3), ")"))

#---
roc_data <- bind_rows(roc_df_imaging, roc_df_combined, 
                      roc_df_cdr, roc_df_bio)


p <- ggplot(roc_data |> arrange(Sensitivity), 
            aes(x = Specificity, y = Sensitivity, color = Model)) +
  geom_line(size = 1.2) + 
  geom_abline(linetype = "dashed", alpha = 0.5) +
  scale_color_manual(values = custom_colors[1:4]) +
  labs(
    title = "ROC Curves: CDR, MRI Imaging, CSF, PET and Demographic Models",
    x = "1 - Specificity", 
    y = "Sensitivity"
  ) +
  theme(
    plot.title = element_text(size = 13, vjust = 1.5),
    legend.position = c(0.7, 0.2), legend.text = element_text(size = 10)
  ) 

p
ggsave("results_baseline_AD-vs-CN/ML_Class.png", 
       p, height = 7, width = 8, dpi = 300)
```

## Neuroimaging Summary and Clinical Relevance

```{r imaging-summary}
cat("NEUROIMAGING KEY FINDINGS:\n")
cat("==========================\n\n")

hip_test <- t.test(Hippocampus_norm ~ Group, data = baseline_data)
cat("Hippocampal Atrophy:\n")
cat("  AD patients show", round((1 - hip_test$estimate[1]/hip_test$estimate[2])*100, 1), 
    "% volume reduction\n")
cat("  Effect size (Cohen's d):", 
    round(cohen.d(baseline_data$Hippocampus_norm ~ baseline_data$Group, na.rm = TRUE)$estimate, 2), "\n")
cat("  p-value:", format.pval(hip_test$p.value), "\n\n")

ento_test <- t.test(Entorhinal_norm ~ Group, data = baseline_data)
cat("Entorhinal Cortex Atrophy:\n")
cat("  AD patients show", round((1 - ento_test$estimate[1]/ento_test$estimate[2])*100, 1), 
    "% volume reduction\n")
cat("  Effect size (Cohen's d):", 
    round(cohen.d(baseline_data$Entorhinal_norm ~ baseline_data$Group, na.rm = TRUE)$estimate, 2), "\n")
cat("  p-value:", format.pval(ento_test$p.value), "\n\n")

vent_test <- t.test(Ventricles_norm ~ Group, data = baseline_data)
cat("Ventricular Enlargement:\n")
cat("  AD patients show", round((vent_test$estimate[1]/vent_test$estimate[2] - 1)*100, 1), 
    "% volume increase\n")
cat("  Effect size (Cohen's d):", 
    round(cohen.d(baseline_data$Ventricles_norm ~ baseline_data$Group, na.rm = TRUE)$estimate, 2), "\n")
cat("  p-value:", format.pval(vent_test$p.value), "\n\n")

cat("Clinical Implications:\n")
cat("- Hippocampal and entorhinal atrophy are hallmark features of AD\n")
cat("- Ventricular enlargement reflects global brain atrophy\n")
cat("- Imaging measures strongly correlate with cognitive performance\n")
cat("- Combined imaging-clinical models achieve excellent diagnostic accuracy\n")
```

# Statistical Modeling

## Logistic Regression: Predictors of AD Diagnosis

```{r logistic-regression}
# Univariate models
uni_vars <- c("AGE", "Male", "PTEDUCAT", "APOE4_1", "APOE4_2", 
              "MMSE_bl", "ADAS11_bl", "FDG_bl", #"MOCA_bl",  "CDRSB_bl", 
              "ABETA_numeric", "TAU_numeric" #,"Hippocampus_norm", "Entorhinal_norm", "VBR"
              )

# data for modeling (complete cases only)
model_data <- baseline_data %>%
  select(Group, uni_vars) %>%
  mutate(AD = ifelse(Group == "AD", 1, 0)) %>%
  na.omit()

model_data_scaled <- model_data %>%
    mutate(across(uni_vars, scale))

cat("Model data dimensions:", dim(model_data_scaled), "\n")

uni_results <- map_dfr(uni_vars, ~{
  formula_str <- paste("AD ~", .x)
  model <- glm(as.formula(formula_str), data = model_data_scaled, family = "binomial")
  tidy(model) %>%
    filter(term == .x) %>%
    mutate(variable = .x)
})

# Multivariate model
multi_model <- glm(AD ~ AGE + Male + PTEDUCAT + APOE4_1 + APOE4_2 + 
                   MMSE_bl + ADAS11_bl + #MOCA_bl +  CDRSB_bl +
                   FDG_bl + ABETA_numeric + TAU_numeric #+ Hippocampus_norm + Entorhinal_norm + VBR
                   ,data = model_data, family = "binomial")

# # here we use stepwise regression (using AIC or BIC criteria) to automatically
# # eliminate predictors and check if the model improves.
# step(multi_model, direction = "both")  # or use "backward" or "forward" based on our need

multi_results <- tidy(multi_model) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    OR = exp(estimate),
    CI_lower = exp(estimate - 1.96 * std.error),
    CI_upper = exp(estimate + 1.96 * std.error)
  )

kable(multi_results %>% 
      select(term, OR, CI_lower, CI_upper, p.value) %>%
      mutate(across(where(is.numeric), round, 3)),
      caption = "Multivariate Logistic Regression Results",
      col.names = c("Variable", "OR", "95% CI Lower", "95% CI Upper", "p-value")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

cat("Model AUC:", pROC::auc(pROC::roc(model_data$AD, predict(multi_model, type = "response"))), "\n")
```

```{r pca_as_predictors_in_regression_model, eval=FALSE, include=FALSE}
# in case many predictors are highly correlated, we might want to apply PCA 
# to reduce dimensionality and remove multicollinearity. PCA will create 
# uncorrelated components that capture the maximum variance in our data.
library(stats)
pca_result <- prcomp(model_data %>%
    select(AGE, Male, PTEDUCAT, APOE4_1, APOE4_2, MMSE_bl, ADAS11_bl, FDG_bl, ABETA_numeric, TAU_numeric, Hippocampus_norm, Entorhinal_norm, VBR),
    scale = TRUE)  # standardize predictors before PCA
summary(pca_result)

```

## Forest Plot of Odds Ratios

```{r forest-plot, fig.height=8, fig.width=10}
forest_data <- multi_results %>%
  mutate(
    Variable_Label = case_when(
      term == "AGE" ~ "Age (per year)",
      term == "Male" ~ "Male gender",
      term == "PTEDUCAT" ~ "Education (per year)",
      term == "APOE4_1" ~ "APOE4 (1 allele)",
      term == "APOE4_2" ~ "APOE4 (2 alleles)",
      term == "MMSE_bl" ~ "MMSE score",
      term == "ADAS11_bl" ~ "ADAS-11 score",
      term == "MOCA_bl" ~ "MOCA score",
      term == "CDRSB_bl" ~ "CDRSB",
      term == "FDG_bl" ~ "FDG-PET",
      term == "ABETA_numeric" ~ "CSF Aβ42",
      term == "TAU_numeric" ~ "CSF Tau",
      TRUE ~ term
    ),
    Significant = ifelse(p.value < 0.05, "Significant", "Non-significant")
  )

ggplot(forest_data, aes(x = reorder(Variable_Label, OR), y = OR)) +
  geom_point(aes(color = Significant), size = 3) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper, color = Significant), 
                width = 0.2) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  coord_flip() +
  scale_color_manual(values = c("Significant" = custom_colors[2], 
                               "Non-significant" = custom_colors[1])) +
  scale_y_log10() +
  labs(title = "Forest Plot: Odds Ratios for AD Diagnosis",
       subtitle = "Multivariate Logistic Regression Model",
       x = "Variable", y = "Odds Ratio (95% CI, log scale)") +
  theme(legend.title = element_blank())
```

# Advanced Visualizations

## Principal Component Analysis

```{r pca-analysis, fig.height=6, fig.width=12}
# PCA on cognitive and biomarker variables
pca_data <- baseline_data %>%
  select(Group, MMSE_bl, ADAS11_bl, ADAS13_bl, LDELTOTAL_BL,
         RAVLT_immediate_bl, FDG_bl, ABETA_numeric, TAU_numeric) %>%
  na.omit()

pca_result <- prcomp(pca_data[,-1], scale = TRUE, center = TRUE)

pca_df <- data.frame(pca_result$x) %>%
  mutate(Group = pca_data$Group)

p1 <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Group)) +
  geom_point(alpha = 0.6, size = 2) +
  stat_ellipse(type = "norm", linetype = 2) +
  scale_color_manual(values = custom_colors[1:2]) +
  labs(title = "Principal Component Analysis",
       subtitle = "Cognitive and Biomarker Variables",
       x = paste0("PC1 (", round(summary(pca_result)$importance[2,1]*100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_result)$importance[2,2]*100, 1), "%)")) +
  theme(legend.position = "none")

scree_data <- data.frame(
  PC = 1:length(pca_result$sdev),
  Variance = pca_result$sdev^2 / sum(pca_result$sdev^2)
)

p2 <- ggplot(scree_data, aes(x = PC, y = Variance)) +
  geom_col(fill = custom_colors[3], alpha = 0.7) +
  geom_line(group = 1) +
  geom_point() +
  labs(title = "Scree Plot", 
       subtitle = "Top 8 Principal Components",
       y = "Proportion of Variance", x = "Principal Component")

grid.arrange(p1, p2, ncol = 2)

ggsave("results_baseline_AD-vs-CN/brain_region_vs_cognitive.png", 
       combined, height = 9, width = 12, dpi = 300)

loadings_df <- data.frame(pca_result$rotation[,1:2]) %>%
  rownames_to_column("Variable") %>%
  mutate(
    Variable_Label = case_when(
      Variable == "MMSE_bl" ~ "MMSE",
      Variable == "ADAS11_bl" ~ "ADAS-11",
      Variable == "ADAS13_bl" ~ "ADAS-13",
      Variable == "LDELTOTAL_BL" ~ "Logical Memory",
      Variable == "RAVLT_immediate_bl" ~ "RAVLT Immediate",
      Variable == "FDG_bl" ~ "FDG-PET",
      Variable == "ABETA_numeric" ~ "CSF Aβ42",
      Variable == "TAU_numeric" ~ "CSF Tau",
      TRUE ~ Variable
    )
  )

p <- ggplot() +
  geom_point(data = pca_df, aes(x = PC1, y = PC2, color = Group), alpha = 0.3) +
  geom_segment(data = loadings_df, 
               aes(x = 0, y = 0, xend = PC1*4, yend = PC2*4),
               arrow = arrow(length = unit(0.3, "cm")), color = "red") +
  stat_ellipse(data = pca_df, aes(x = PC1, y = PC2, color = Group), 
               type = "norm", linetype = 2, show.legend = FALSE) +
  geom_text_repel(data = loadings_df, 
            aes(x = PC1*4.5, y = PC2*4.5, label = Variable_Label),
            size = 4, alpha = 0.7, segment.size = .25, segment.alpha = .8,
            force = 1, color = "red", fontface = "bold",
            position = position_dodge(1)) +
  scale_color_manual(values = custom_colors[1:2]) +
  labs(title = "PCA Biplot", 
       subtitle = "Variables (red) and Individuals (colored by group)")
p

full_layout <- arrangeGrob(grobs = list(p, arrangeGrob(p1, p2, ncol = 2)),
  layout_matrix = rbind(c(1, 1), c(2, 2)))

combined <- arrangeGrob(p, p2, ncol = 2)

ggsave("results_baseline_AD-vs-CN/pca_analysis.png",
       plot = combined, width = 11, height = 6, dpi = 300)
```

```{r pca-analysis, eval=FALSE, include=FALSE}
# PCA on cognitive and biomarker variables
pca_data <- baseline_data %>%
  select(Group, MMSE_bl, ADAS11_bl, ADAS13_bl, LDELTOTAL_BL,
         RAVLT_immediate_bl, FDG_bl, ABETA_numeric, TAU_numeric) %>%
  na.omit()

pca_result <- prcomp(pca_data[,-1], scale. = TRUE, center = TRUE)

# Data for PCA scores and loadings
pca_df <- data.frame(pca_result$x) %>%
  mutate(Group = pca_data$Group)

loadings_df <- data.frame(pca_result$rotation[,1:2]) %>%
  rownames_to_column("Variable") %>%
  mutate(
    Variable_Label = case_when(
      Variable == "MMSE_bl" ~ "MMSE",
      Variable == "ADAS11_bl" ~ "ADAS-11",
      Variable == "ADAS13_bl" ~ "ADAS-13",
      Variable == "LDELTOTAL_BL" ~ "Logical Memory",
      Variable == "RAVLT_immediate_bl" ~ "RAVLT Immediate",
      Variable == "FDG_bl" ~ "FDG-PET",
      Variable == "ABETA_numeric" ~ "CSF Aβ42",
      Variable == "TAU_numeric" ~ "CSF Tau",
      TRUE ~ Variable
    )
  )

# PCA biplot
p1 <- ggplot() +
  geom_point(data = pca_df, aes(x = PC1, y = PC2, color = Group), alpha = 0.8, size = 2) +
  stat_ellipse(data = pca_df, aes(x = PC1, y = PC2, color = Group), 
               type = "norm", linetype = 2, show.legend = FALSE) +
  geom_segment(data = loadings_df, 
               aes(x = 0, y = 0, xend = PC1 * 5, yend = PC2 * 5),
               arrow = arrow(length = unit(0.3, "cm")), color = "red") +
  geom_text_repel(data = loadings_df, 
                  aes(x = PC1 * 5.5, y = PC2 * 5.5, label = Variable_Label),
                  size = 4, alpha = 0.8, color = "red", fontface = "bold") +
  scale_color_manual(values = custom_colors[1:2]) +
  labs(
    title = "PCA Biplot",
    subtitle = "Cognitive and Biomarker Variables",
    x = paste0("PC1 (", round(summary(pca_result)$importance[2,1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_result)$importance[2,2] * 100, 1), "%)")
  ) +
  theme_minimal()

# Scree plot
scree_data <- data.frame(
  PC = 1:length(pca_result$sdev),
  Variance = pca_result$sdev^2 / sum(pca_result$sdev^2)
)

p2 <- ggplot(scree_data, aes(x = PC, y = Variance)) +
  geom_col(fill = custom_colors[3], alpha = 0.7) +
  geom_line(group = 1) +
  geom_point() +
  labs(
    title = "Scree Plot", 
    subtitle = "Variance Explained by Principal Components",
    y = "Proportion of Variance", x = "Principal Component"
  ) +
  theme_minimal()

# Combine and save
combined <- gridExtra::arrangeGrob(p1, p2, ncol = 2)

ggsave("results_baseline_AD-vs-CN/pca_analysis.png",
       plot = combined, width = 12, height = 6.5, dpi = 300)

```


# Correlations of all measurements

```{r}
# we do correlation between brain measures and cognitive scores
all_data <- baseline_data %>%
  select("Group", cog_vars, bio_vars, imaging_vars) %>%
  mutate(across(everything(), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .))) %>%
  select(where(~ mean(is.na(.)) < 0.3))
 
all_data <- all_data %>%
  rename(
    MMSE = MMSE_bl,
    `ADAS-11` = ADAS11_bl,
    `ADAS-13` = ADAS13_bl,
    `ADAS-Q4` = ADASQ4_bl,
    `Logical Memory` = LDELTOTAL_BL,
    `RAVLT Immediate` = RAVLT_immediate_bl,
    `RAVLT Learning` = RAVLT_learning_bl,
    `RAVLT Forgetting` = RAVLT_forgetting_bl,
    MOCA = MOCA_bl,
    `CDR-SB` = CDRSB_bl,
    `Digit Symbol Substitution` = DIGITSCOR_bl,
    FAQ = FAQ_bl,
    `FDG-PET` = FDG_bl,
    `Amyloid PET` = AV45_bl,
    "CSF A\u03B242" = ABETA_numeric,
    `CSF Tau` = TAU_numeric,
    Ventricles = Ventricles_norm,
    `Intracranial Volume ` = ICV_bl,
    `Ventricular-to-Brain Ratio ` = VBR,
    Hippocampus = Hippocampus_norm,
    `Whole Brain` = WholeBrain_norm,
    `Entorhinal Cortex` = Entorhinal_norm,
    `Fusiform Gyrus` = Fusiform_norm,
    `Middle Temporal` = MidTemp_norm
  )

# corr. matrix
cor_matrix <- cor(all_data %>% select(-Group), use = "pairwise.complete.obs")

png("results_baseline_AD-vs-CN/all_corrplot.png", width = 3100, height = 2400, res = 150)
corrplot(cor_matrix, method = "color", order = "hclust", 
         addCoef.col = "black", tl.col = "black", tl.srt = 45, 
         number.cex = 1.2, tl.cex = 1.4, sig.level = 0.01, cex.main = 2.1,  
         insig = "blank", diag = FALSE, cl.cex = 1.4,
         title = "Correlations: Brain Structure, Cognitive Measures, Imaging and CSF Biomarkers"
         ,mar = c(1, 1, 6, 1)
         )    # margin: bottom, left, top, right
dev.off()
```

# Summary and Conclusions

```{r summary-stats}
cat("Key Findings:\n")
cat("=============\n\n")

n_ad <- sum(baseline_data$Group == "AD", na.rm = TRUE)
n_cn <- sum(baseline_data$Group == "CN", na.rm = TRUE)
cat("Sample Sizes: AD =", n_ad, ", CN =", n_cn, "\n\n")

age_test <- t.test(AGE ~ Group, data = baseline_data)
cat("Age: AD patients are significantly older than CN controls\n")
cat("  Mean difference:", round(age_test$estimate[1] - age_test$estimate[2], 1), "years\n")
cat("  p-value:", format.pval(age_test$p.value, digits = 3), "\n\n")

mmse_test <- t.test(MMSE_bl ~ Group, data = baseline_data)
cat("MMSE: AD patients have significantly lower scores\n")
cat("  Mean difference:", round(mmse_test$estimate[2] - mmse_test$estimate[1], 1), "points\n")
cat("  p-value:", format.pval(mmse_test$p.value, digits = 3), "\n\n")

apoe_test <- chisq.test(table(baseline_data$Group, baseline_data$APOE4))
cat("APOE4: Significantly associated with AD diagnosis\n")
cat("  Chi-square p-value:", format.pval(apoe_test$p.value, digits = 3), "\n")
```

## Key Statistical Differences

The analysis reveals significant differences between AD patients and CN
controls across multiple domains:

**Demographics:** - AD patients are significantly older than CN
controls - Higher proportion of APOE4 carriers in AD group - No
significant gender differences

**Cognitive Function:** - AD patients show severe impairment on all
cognitive measures - Largest effect sizes observed for ADAS scores and
memory tests - MMSE scores clearly differentiate the groups

**Biomarkers:** - Lower CSF Aβ42 levels in AD patients - Higher CSF Tau
levels in AD patients\
- Reduced FDG-PET metabolism in AD group

**Clinical Implications:** The multivariate model identifies key
predictors of AD diagnosis, with cognitive measures and biomarkers
showing the strongest associations. These findings support the use of
combined cognitive and biomarker assessments for AD diagnosis and
monitoring.

------------------------------------------------------------------------

*Analysis completed on `r Sys.Date()`* *Data source: ADNI (Alzheimer's
Disease Neuroimaging Initiative)*
