---
title: "Elastic Net regularization (mixed L1 + L2)"
author: "Mukesha et al."
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    self_contained: true
    code_folding: hide
  word_document:
    toc: true
  pdf_document:
    toc: true
    latex_engine: xelatex
bibliography: AD_vs_CN_references.bib
editor_options:
  markdown:
    wrap: 72
abstract: |
  This report compares LASSO (α = 1), Ridge (α = 0), and Elastic Net (α ∈
  [0,1]) regularization methods on metabolomics data distinguishing
  Alzheimer’s disease (AD) from cognitively normal controls (CN). The analysis
  uses glmnet with repeated cross-validation, evaluates tuning parameters
  (alpha, lambda), generates coefficient paths, identifies selected
  metabolites, and produces publication-quality visualizations.
---

in this version, I:

* remove redundant code
* use a clear training/validation split
* eliminate your complex seed-handling (not needed)
* use a simple `tuneGrid` so you control α directly
* parallelize safely
* produce variable selection and plots cleanly

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(caret)
library(glmnet)
library(doParallel)
library(dplyr)
library(ggplot2)
library(tibble)
library(kableExtra)
library(tidyr)
library(readr)
```

# Load and prepare data

```{r load}
data <- read.csv("03_EffectsAdjusted.csv")
data$Allgr <- factor(data$Allgr, levels = c("CN", "AD"))

X <- data[, 8:ncol(data)]  
y <- data$Allgr
```

# Train/test split

```{r split}
set.seed(123)
train_index <- createDataPartition(y, p = 0.7, list = FALSE)

trainX <- X[train_index, ]
trainY <- y[train_index]

testX  <- X[-train_index, ]
testY  <- y[-train_index]
```

# Fit Ridge, Elastic Net, and LASSO

```{r}
set.seed(2024)

alpha_values <- c(0, 0.1, 0.5, 0.9, 1)   # Ridge → Elastic Net → LASSO
lambda_seq <- exp(seq(log(0.0001), log(1), length.out = 50))

tune <- expand.grid(alpha = alpha_values, lambda = lambda_seq)

cl <- makeCluster(detectCores() - 2)
registerDoParallel(cl)
ctrl <- trainControl(
    method = "repeatedcv",
    number = 10, repeats = 5,
    summaryFunction = twoClassSummary,
    classProbs = TRUE,
    savePredictions = TRUE
)

fit <- train(
    x = trainX,
    y = trainY,
    method = "glmnet",
    metric = "ROC",
    trControl = ctrl,
    tuneGrid = tune,
    preProcess = c("center", "scale", "medianImpute")
)
```


# Comparison Table (Ridge vs EN vs LASSO)**

```{r}
perf_table <- fit$results %>%
    group_by(alpha) %>%
    summarize(
        Best_ROC = max(ROC),
        Best_lambda = lambda[which.max(ROC)]
    )

perf_table %>%
    kable("html", caption = "Performance Comparison Across Regularization Types") %>%
    kable_styling(full_width = FALSE)
```

# Coefficient Path Plot (LASSO path)**

We use `glmnet` directly for path plotting:

```{r}
lasso_path <- glmnet(
    as.matrix(trainX),
    trainY,
    alpha = 1,
    family = "binomial",
    standardize = TRUE
)

stopCluster(cl)
registerDoSEQ()

png("LASSO_path_plot.png", width = 1500, height = 1200, res = 180)
plot(lasso_path, xvar = "lambda", label = TRUE)
dev.off()
```

# Top features

```{r}
best_alpha  <- fit$bestTune$alpha
best_lambda <- fit$bestTune$lambda

coef_mat <- coef(fit$finalModel, s = best_lambda)
coef_df <- data.frame(
    Feature = rownames(coef_mat),
    Coef = as.numeric(coef_mat)
) %>%
    filter(Feature != "(Intercept)", Coef != 0) %>%
    arrange(desc(abs(Coef))) %>%
    head(50)   # top 50 for readability

p <- ggplot(coef_df, aes(x = reorder(Feature, abs(Coef)), y = Coef, fill = Coef > 0)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    scale_fill_manual(values = c("#D55E00", "#0072B2")) +
    theme_bw(base_size = 13) +
    labs(
        title = "Top 50 Features Selected by Elastic Net",
        subtitle = paste("alpha =", best_alpha, "| lambda =", round(best_lambda, 5)),
        x = "Metabolites",
        y = "Coefficient"
    )
p
ggsave("Top50_EN_features_publication.png", p, width = 8, height = 10, dpi = 300)
```

# Export Selected Metabolites + Coefficients

```{r}
selected_features <- coef_df

write.csv(selected_features, "Selected_Metabolites_EN.csv", row.names = FALSE)
saveRDS(selected_features, "Selected_Metabolites_EN.rds")
```

# Save text summary

```{r}
sink("feature_selection_resport.txt")

cat("===== FEATURE SELECTION REPORT =====\n")
cat("Best alpha:", best_alpha, "\n")
cat("Best lambda:", best_lambda, "\n\n")

cat("=== Number of Selected Features ===\n")
cat(nrow(selected_features), "\n\n")

cat("=== Top 10 Features ===\n")
print(head(selected_features, 10))

cat("\n=== Performance Summary ===\n")
print(perf_table)

sink()
```
```{r}
cat("===== FEATURE SELECTION REPORT =====\n")
cat("Best alpha:", best_alpha, "\n")
cat("Best lambda:", best_lambda, "\n\n")

cat("=== Number of Selected Features ===\n")
cat(nrow(selected_features), "\n\n")

cat("=== Top 10 Features ===\n")
print(head(selected_features, 10))

cat("\n=== Performance Summary ===\n")
print(perf_table)
```

