---
title: "Biocrates data analysis report"
author: "Firalis/Lodiag"
date: "`2024-07-30`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    self_contained: true
  word_document:
    toc: true
  pdf_document:
    toc: true
always_allow_html: true
editor_options:
  markdown:
    wrap: 72
bibliography: references.bib
---

```{r libraries}
library(knitr)
library(ADNIMetabo)

library(dplyr)
library(ggplot2)
library(ggpubr)
library(gridExtra)
library(cowplot)
library(ggrepel)

library(FactoMineR)
library(data.table)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
                      collapse = TRUE, comment = "#>")

# knitr::opts_chunk$set(echo = TRUE)
```

## **Background**

Alzheimer's disease (AD) poses a significant global health challenge and
current diagnostic methods, particularly those involving cerebro-spinal
fluid (CSF), are invasive and often lack the sensitivity needed for
early detection. Studies has highlighted the potential of blood-based
biomarkers, particularly lipidomic profiles, to identify pre-clinical AD
effectively [@mapstone2014]. As an example, Lipidomic studies have
identified specific lipids in serum that can predict progression to mild
cognitive impairment or AD with over 90% accuracy within 2-3 years
[@oka2024]. In addition, blood tests measuring phosphorylated tau and
amyloid ratios have demonstrated high diagnostic
accuracy[@palmqvist2024]. Metabolomic analyses have further revealed key
metabolic alterations associated with AD, including reduced plasmalogen
lipids and increased gangliosides, suggesting that a combination of
lipidomic and metabolomic data could significantly enhance early
detection strategies[@oka2024]. However, the practical application of
these biomarkers in diverse populations and clinical settings still
requires extensive validation[@oka2024]. In this study, we aim to
contribute by providing a new insight into metabolomic biomarkers using
Biocrates panels, evaluating metabolic alterations that may be
associated to AD.

------------------------------------------------------------------------

> | Source code and results

# Data preparation and quality control

## Missing data analysis

```{r missing data}
data0 <- data.table::fread(input = "../data-raw/adni_serum_ADvsHC.csv")

data0[, Gender := toupper(Gender)]
data0[Gender %in% c("1", "M"), Gender := "MALE"]
data0[Gender %in% c("2", "F"), Gender := "FEMALE"]
data0[, `:=`(Gender = factor(Gender))]

biomarkers <- colnames(data0)[8:dim(data0)[2]]
data0[, `:=`(Gender = factor(Gender))]
data0$Allgr <- as.factor(data0$Allgr)
data0$Allgr <- relevel(data0$Allgr, ref = "AD")
data0$Allgr <- relevel(data0$Allgr, ref = "CN")
metadata <- c("Gender", "Allgr", "Age" )
```


# Descriptive Statistics

```{r descrptive_plot, fig.width=12, fig.height=10, fig.align='center', out.width='100%'}
color_palette <- c("#1f78b4", "#33a02c")
color_gender <- c("#9f9f1a", "#6a3d1a")

data0 <- as.data.table(data0)

gender_cohort <- data0 |>
  dplyr::group_by(Allgr, Gender) |>
  dplyr::summarise(patients = dplyr::n()) |>
  dplyr::mutate(group_proportion = scales::percent(patients / sum(patients)))

p2 <- ggplot(gender_cohort, aes(x = Allgr, y = patients, fill = Gender)) +
  geom_bar(stat = "identity", position = "stack",width = 0.5) +
  geom_text(aes(label = paste0(patients, "\n(", group_proportion, ")")), 
            position = position_stack(vjust = 0.5), 
            size = 3, color = "white") +
  scale_fill_manual(values = color_gender) +
  labs(title = "Sex distribution across groups",
       x = "Groups", y = "Number of Patients", fill = "Sex") +
  theme_bw() +
   theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5))

p3 <- ggplot(data0, aes(x = Allgr, y = Age)) +
  geom_violin(aes(fill = Gender), trim = FALSE, color = "black", 
              alpha = 0.6, position = position_dodge(1)) +
  geom_boxplot(width = 0.2, position = position_dodge(1), color = "black", 
               aes(fill = Gender), outlier.shape = NA) +
  scale_fill_manual(values = color_gender) +
  scale_color_manual(values = color_gender) +
  labs(title = "Age distribution across groups", 
       x = "Groups", y = "Age") +
  theme_bw() + ylim(20, 100) + guides(fill = "none") +
  theme(legend.position = "right", plot.title = element_text(hjust = 0.5))


# 3. Age distribution across groups (violin plot)
p3 <- ggplot(data0, aes(x = Allgr, y = Age)) +
  geom_violin(aes(fill = Gender), trim = FALSE, color = "black", 
              alpha = 0.6, position = position_dodge(1)) +
  geom_boxplot(width = 0.2, position = position_dodge(1), color = "black", 
               aes(fill = Gender), outlier.shape = NA) +
  scale_fill_manual(values = color_gender) +
  scale_color_manual(values = color_gender) +
  labs(title = "Age distribution across groups", 
       x = "Groups", y = "Age") +
  theme_bw() + ylim(20, 100) + guides(fill = "none") +
  theme(legend.position = "right", plot.title = element_text(hjust = 0.5))

# 4. APOE genotype distribution across sub-groups (violin plot)
p4 <- data0 |>
  dplyr::group_by(Allgr, Gender, APOE4) |>
  dplyr::summarise(n = dplyr::n()) |>
  ggplot(aes(x = Allgr, y = n, fill = Gender)) + 
  scale_fill_manual(values = color_gender) +
  geom_col(position = "dodge") + 
  facet_wrap(~APOE4) +
  labs(title = "Distribution by Diagnosis, Sex, and APOE4",
       x = "Diagnosis", y = "Count") +
    theme_bw() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.05),
        legend.text = element_text(size = 11),
        strip.text = element_text(size = 12))

combined_plot <- gridExtra::grid.arrange(
  ggpubr::ggarrange(p3, p2, common.legend = TRUE,
            legend="right", ncol = 2), 
  p4, nrow = 2)

combined_plot1 <- ggpubr::as_ggplot(combined_plot) +                        
    cowplot::draw_plot_label(label = c("A)", "B)", "C)"), size = 15,
                    x = c(0.03, 0.47, 0.03), y = c(1, 1, 0.5))
ggsave("pictures/dea_approfond.png", combined_plot1, 
       width = 7.5, height = 7.5, dpi = 300)

```


```{r descriptive_stats_by_age}
summary <- data0 |>
  dplyr::group_by(Allgr) |>
  dplyr::summarise(
    n = dplyr::n(),
    Age_mean = round(mean(Age, na.rm = TRUE), 1),
    Age_sd = round(sd(Age, na.rm = TRUE), 1),
    Age_min = min(Age, na.rm = TRUE),
    Age_max = max(Age, na.rm = TRUE)
  )
DT::datatable(summary, caption = "Descriptive statistics for age by subgroup", 
              colnames = c("Allgr", "N", "Mean", "SD", "Min", "Max")) 
```

# PCA results

## Extremes and outliers

```{r pca1}
data2 <- data0 # missing data was already handled before
biomarkers <- data2[, 8:ncol(data2)]
coul = rainbow(3)
acp = FactoMineR::PCA(biomarkers)

plot(as.numeric(acp$ind$coord[,1]), as.numeric(acp$ind$coord[,2]),
     col = coul[as.factor(data2$Allgr)], main = "Representation of sample by group",
     xlab = paste("PC1: ", round(acp$eig[1,2],2), "%", sep = ""),
     ylab = paste("PC2: ",round(acp$eig[2,2],2),"%",sep = ""),pch = 16)
abline(v = 0,h = 0,lty = 3)
legend("topleft",legend = levels(as.factor(data2$Allgr)),pch = 16,col = coul, bty = 'n')
coul = rainbow(3)

plot(as.numeric(acp$ind$coord[,1]), as.numeric(acp$ind$coord[,2]),
     col = coul[as.factor(data2$APOE4)], main = "Representation of sample by APOE4", 
     xlab = paste("PC1: ",round(acp$eig[1,2],2),"%",sep = ""), 
     ylab = paste("PC2: ", round(acp$eig[2,2],2),"%",sep = ""), pch = 16)
abline(v = 0,h = 0,lty = 3)
legend("topleft", legend = levels(as.factor(data2$APOE4)), pch = 16, col = coul, bty = 'n')
coul = rainbow(2)

plot(as.numeric(acp$ind$coord[,1]), as.numeric(acp$ind$coord[,2]), 
     col = coul[as.factor(data2$Gender)], main = "Representation of sample by sex", 
     xlab = paste("PC1: ",round(acp$eig[1,2],2), "%", sep = ""), 
     ylab = paste("PC2: ",round(acp$eig[2,2],2),"%", sep = ""), pch = 16)
abline(v = 0, h = 0, lty = 3)
legend("topleft", legend = levels(as.factor(data2$Gender)), pch = 16, col = coul,bty = 'n')
```


# Multivariave ANOVA to evaluate Batch, Age and Gender effect

```{r anova, messAge=FALSE}

data2$Allgr = as.factor(data2$Allgr)
data2$Gender = as.factor(data2$Gender)
data2$Plate = as.factor(data2$Plate)

# Remove rows with missing values
# data2 <- na.omit(data2)

data2 <- as.data.frame(data2)
test = NULL
for (i in 7:ncol(data2)) {
  y <- data2[[i]]
  if (!is.numeric(y) || anyNA(y)) next
  test = rbind(test, c(colnames(data2)[i], summary(
    aov(
      data2[, i] ~ data2$Plate + data2$Gender * data2$Age + data2$Allgr,
      na.rm = T
    )
  )[[1]][["Pr(>F)"]]))
}
test = test[,-7]
colnames(test) = c("Biomarker", "Batch", "Gender", "Age", "Group", "Gender:Age")
test = as.data.table(test)

test_long = melt(test, id.vars = c("Biomarker"))
test_long$value = as.numeric(test_long$value)
test_long$pvalue_adjusted = p.adjust(test_long$value, method = "fdr")
#test_wide <- data.table::dcast(test_long, Biomarker ~ variable, 
#                   value.var =  c("value","pvalue_adjusted"))

test_wide <- data.table::dcast(
  test_long |> as.data.table(),
  Biomarker ~ variable,
  value.var = c("value", "pvalue_adjusted")
)

```

```{r anovagraphs, messAge=FALSE}
anova_results <- test_wide

group_plot <- ggplot(anova_results, aes(value_Group)) +
  geom_histogram(fill = "#56B4E9", alpha = 0.8) +
  geom_vline(xintercept = c(0.05), linetype = "dashed", color = "red") +
  ggtitle("group p-values distribution") +
  ylab("Count") +
  xlab("p-values") +
  theme_bw()

batch_plot <- ggplot(anova_results, aes(value_Batch)) +
  geom_histogram(fill = "#56B4E9", alpha = 0.8) +
  geom_vline(xintercept = c(0.05), linetype = "dashed", color = "red") +
  ggtitle("batch p-values distribution") +
  ylab("Count") +
  xlab("p-values") +
  theme_bw()

Age_plot <- ggplot(anova_results, aes(value_Age)) +
  geom_histogram(fill = "#56B4E9", alpha = 0.8) +
  geom_vline(xintercept = c(0.05), linetype = "dashed", color = "red") +
  ggtitle("age p-values distribution") +
  ylab("Count") +
  xlab("p-values") +
  theme_bw()

gender_plot <- ggplot(anova_results, aes(value_Gender)) +
  geom_histogram(fill = "#56B4E9", alpha = 0.8) +
  geom_vline(xintercept = c(0.05), linetype = "dashed", color = "red") +
  ggtitle("sex p-values distribution") +
  ylab("Count") +
  xlab("p-values") +
  theme_bw()

Age_gender_plot <- ggplot(anova_results, aes(`value_Gender:Age`)) +
  geom_histogram(fill = "#56B4E9", alpha = 0.8) +
  geom_vline(xintercept = c(0.05), linetype = "dashed", color = "red") +
  ggtitle("age:sex p-values distribution") +
  ylab("Count") +
  xlab("p-values") +
  theme_bw()

adj_group_plot <- ggplot(anova_results, aes(pvalue_adjusted_Group)) +
  geom_histogram(fill = "#56B4E9", alpha = 0.8) +
  geom_vline(xintercept = c(0.05), linetype = "dashed", color = "red") +
  ggtitle("group q-values distribution") +
  ylab("Count") +
  xlab("q-values") +
  theme_bw()

adj_batch_plot <- ggplot(anova_results, aes(pvalue_adjusted_Batch)) +
  geom_histogram(fill = "#56B4E9", alpha = 0.8) +
  geom_vline(xintercept = c(0.05), linetype = "dashed", color = "red") +
  ggtitle("batch q-values distribution") +
  ylab("Count") +
  xlab("q-values") +
  theme_bw()

adj_Age_plot <- ggplot(anova_results, aes(pvalue_adjusted_Age)) +
  geom_histogram(fill = "#56B4E9", alpha = 0.8) +
  geom_vline(xintercept = c(0.05), linetype = "dashed", color = "red") +
  ggtitle("age q-values distribution") +
  ylab("Count") +
  xlab("q-values") +
  theme_bw()

adj_gender_plot <- ggplot(anova_results, aes(pvalue_adjusted_Gender)) +
  geom_histogram(fill = "#56B4E9", alpha = 0.8) +
  geom_vline(xintercept = c(0.05), linetype = "dashed", color = "red") +
  ggtitle("sex q-values distribution") +
  ylab("Count") +
  xlab("q-values") +
  theme_bw()

adj_Age_gender_plot <- ggplot(anova_results, aes(`pvalue_adjusted_Gender:Age`)) +
  geom_histogram(fill = "#56B4E9", alpha = 0.8) +
  geom_vline(xintercept = c(0.05), linetype = "dashed", color = "red") +
  ggtitle("age:sex q-values distribution") +
  ylab("Count") +
  xlab("q-values") +
  theme_bw()

gridExtra::grid.arrange(group_plot, adj_group_plot, batch_plot, adj_batch_plot, nrow = 2, ncol = 2) 

gridExtra::grid.arrange(Age_plot, adj_Age_plot,gender_plot, adj_gender_plot, nrow = 2, ncol = 2)

```

## Multivariate ANOVA results

```{r anovatables, messAge=FALSE}
anova_results <- as.data.frame(anova_results)
anova_results[,-1] <- round(anova_results[,-1],3)
colnames(anova_results)[2]  <- "p_Plate" 
colnames(anova_results)[3]  <- "p_Gender" 
colnames(anova_results)[4] <- "p_Age"
colnames(anova_results)[5] <- "p_Group"
colnames(anova_results)[6] <- "p_Gender:Age"
colnames(anova_results)[7]  <- "q_Plate" 
colnames(anova_results)[8]  <- "q_Gender" 
colnames(anova_results)[9] <- "q_Age"
colnames(anova_results)[10] <- "q_Group"
colnames(anova_results)[11] <- "q_Gender:Age"
DT::datatable(anova_results)
```


## Data adjustment

Here we are going to perform a data adjustment process across multiple
variables in a dataset, by calculating predictions based on these models
and subtracts certain predicted effects (likely the main effects of the
predictors) from the original data. The process involves linear
regression modeling and selective removal of predictor influences from
multiple variables. The process removes the influence of these predictor
variables from each of the analyzed columns, which could be interpreted
as a form of batch correction or covariate adjustment. This technique is
often used to account for known sources of variation and isolate the
effects of interest.

The input dataset, referred to as data2, contains multiple columns
including:

-   Plate
-   Gender
-   Age
-   Cohort
-   Various measurement variables (from column 8 onwards)

Purpose and Implications

-   This process likely serves as a form of batch correction or
    covariate adjustment.
-   It aims to remove the effects of Plate, Gender, Age, and Cohort from
    each measurement variable.
-   The adjustment may help isolate effects of interest by accounting
    for known sources of variation.
    
### Refining the Adjustment Model

Check p-values from ANOVA (test_wide)

```{r}
significant_covariates = test_wide[, lapply(.SD, function(x)
  sum(x < 0.05, na.rm = TRUE)), .SDcols = patterns("pvalue_adjusted")]
print(significant_covariates)
```

Adjust the Linear Model Based on ANOVA

```{r adjusteddata}
data0_1 = data2[,1:7]

for (i in 8:ncol(data2)) {
  model = lm(data2[, i] ~ data2$Plate + data2$Gender * data2$Age + data2$Allgr)
  coeff = as.matrix(coef(summary(model), keep.rownames = "coefficient"))
  pred = stats::predict(model, data2, type = "terms")
  data0_1 = cbind(data0_1, data2[, i] - (pred[, 1] + pred[, 2] + pred[, 3] + 
                                           pred[, 5]))
}
colnames(data0_1)[8:ncol(data0_1)] = colnames(data2)[8:ncol(data0_1)]

write.csv(data0_1,"03_EffectsAdjusted.csv",row.names = F)
```

## PCA after adjustment

```{r afeter_effect_removal_by_ml}
biomarkers <- data0_1[, 8:ncol(data0_1)]
coul = rainbow(59)
acp = FactoMineR::PCA(biomarkers)
plot(
  as.numeric(acp$ind$coord[, 1]), as.numeric(acp$ind$coord[, 2]),
  col = coul[as.factor(data0_1$Plate)],
  main = "Representation of sample after batch effect correction",
  xlab = paste("Dim1: ", round(acp$eig[1, 2], 2), "%", sep = ""),
  ylab = paste("Dim2: ", round(acp$eig[2, 2], 2), "%", sep = ""), pch = 16)
abline(v = 0, h = 0, lty = 3)
legend("topright",legend = levels(as.factor(data0_1$Plate)), pch = 16, col = coul, 
       bty = 'n',  # No box around the legend
       title = "Plate", 
       cex = 0.5,  # Adjust text size
       xpd = TRUE,  # Allow legend to go outside plot area if needed
       inset = c(-0.17, -0.1)  # Adjust the legend's position to the right
)
```
```{r 3dpca_rgl, eval=TRUE, include=FALSE, fig.width=6, fig.height=5, fig.align='center', out.width='100%'}
first_two_components <- data.frame(
  PC1 = as.numeric(acp$ind$coord[, 1]), 
  PC2 = as.numeric(acp$ind$coord[, 2]), 
  Group = data0_1$Allgr,
  Sex = data0_1$Gender)

coul = rainbow(3)
colors <- coul[as.factor(first_two_components$Group |> unique())]

shapes <- c(15, 16)

pca_plot2d <- ggplot(first_two_components, aes(x = PC1, y = PC2, color = Group 
                                 # ,shape = Sex
                                 )) +
    geom_point(alpha = 0.8, size = 3, 
               # shape = shapes[as.numeric(first_two_components$Gender)],
               show.legend = TRUE) +
    # labs(title = "Metabolomics data projected onto the first 2 Principal Components") +
    scale_color_manual(values = colors) +
    scale_shape_manual(values = shapes) +
    theme_classic() + ylim(-20, 20) + xlim(-25, 25) +
    ylab(label =  paste("PC1: ", round(acp$eig[1, 2], 2), "%", sep = "")) + 
    xlab(label =  paste("PC2: ", round(acp$eig[2, 2], 2), "%", sep = "")) +
    theme(plot.title = element_text(hjust = 0.5, vjust = 0.9),
          legend.position = c(0.9, 0.8),
          legend.spacing.y = unit(0.2, 'cm'),
          axis.text = element_text(size = 12),
          axis.title = element_text(size = 13),
          legend.background =
              element_rect(fill = "seashell", size = 0.5,
                           linetype = "solid", colour = "lightblue3"),
          legend.title = element_text(size = 11),
          legend.text = element_text(size = 10)
          ) 

pca_summary <- acp$eig |> data.frame() |>
  dplyr::slice_head(n = 4) |>
  dplyr::rename("CumulativeExplainedVariance" =
                  "cumulative.percentage.of.variance",
                "ExplainedVariance" = "percentage.of.variance") |>
  dplyr::mutate("Component" = acp$eig |> data.frame() |>
                  dplyr::slice_head(n = 4) |> rownames()) |>
  dplyr::mutate(across('Component', stringr::str_replace, 'comp ', ''))

explained_variance_plot <- ggplot(pca_summary, aes(x = Component)) +
  geom_bar(aes(y = ExplainedVariance, fill = "Explained Variance"), 
           stat = "identity", color = "black") +
  geom_text(aes(y = ExplainedVariance, 
                label = ifelse(Component != 1, paste0(round(ExplainedVariance, 2), "%"), "")), 
            vjust = -0.5, size = 5, color = "#000000") +
  geom_line(aes(y = CumulativeExplainedVariance), color = "orange", size = 1, group = 1) +
  geom_point(aes(y = CumulativeExplainedVariance, color = "Cumulative Explained Variance"), size = 3) +
  geom_text(aes(y = CumulativeExplainedVariance, 
                label = paste0(round(CumulativeExplainedVariance, 2), "%")),
            vjust = -0.5, size = 5, color = "#000000") +
  scale_y_continuous(name = "Explained Variance (%)", limits = c(0, 101)) +
  labs(x = "Principal Component", fill = "Legend", color = "Legend") +
  scale_fill_manual(values = c("Explained Variance" = "steelblue")) +
  scale_color_manual(values = c("Cumulative Explained Variance" = "orange4")) +
  theme_classic() +
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 13),
        legend.title = element_blank(),
        legend.text = element_text(size = 11),
        legend.position = c(0.52, 0.88),
        legend.spacing.y = unit(0.01, 'cm'), 
        legend.justification = c(0, 1),  
        legend.direction = "vertical") 

print(combined_plot)


# grid.arrange(
#   pca_plot2d,
#   plot_explained_variance,
#   plot_cumulative_explained_variance, 
#   ncol = 3,  nrow = 1,
#   top = textGrob("PCA Variance Explained", 
#                  gp = gpar(fontsize = 13, col = "black"))
# )
pca_results <- gridExtra::arrangeGrob(pca_plot2d, explained_variance_plot, ncol = 2)

pca_results1 <- ggpubr::as_ggplot(pca_results) +                                
    cowplot::draw_plot_label(label = c("A", "B"), size = 20,
                    x = c(0.06, 0.57), y = c(1, 1))   

ggsave("pictures/PCA with Variance Explained.tiff", pca_results1, 
       width = 13, height = 6, device = 'tiff', dpi = 1200)
pca_results1
```

# Comparison of AD vs CN

## Volcano plot for p values

Student t test the p-values and the q-values FC distributions are
displayed below.

```{r, messAge=FALSE}
CN = data0_1[which(data0_1$Allgr == "CN"), ]
AD = data0_1[which(data0_1$Allgr == "AD"), ]
n = ncol(CN)
dat = rbind(CN, AD)
test = ADNIMetabo::student_fdr(8, n, AD, CN, F, 2, "AD vs CN")
FC <- as.numeric(test[, 8])
Pvalue <- as.numeric(test[, 6])
threshold = 1.25
logFC <- log2(FC)
wh <- which(Pvalue == 0)
Pvalue[wh] <- 0.000001
logpval <- -log10(Pvalue)
res <- data.frame(cbind(logFC, logpval))
  
with(res, plot(logFC, logpval, pch = 20, ylab = "p_value (in -log10)", 
               xlab = "Fold change (in log2)"),cex.axis = 3,cex.lab = 3)
with(subset(res, logpval < -log10(0.05) ), 
     points(logFC, logpval, pch = 20, col = "black"))
with(subset(res, logFC > log2(1) & logpval > -log10(0.05)), 
     points(logFC,  logpval, pch = 20, col = "lightcoral"))
with(subset(res, logFC < log2(1) & logpval > -log10(0.05)), 
     points(logFC,  logpval, pch = 20, col = "lightblue"))
with(subset(res, logFC > log2(threshold) & logpval > -log10(0.05)), 
     points(logFC,  logpval, pch = 20, col = "red"))
with(subset(res, logFC < log2(1/threshold) & logpval > -log10(0.05)), 
     points(logFC,  logpval, pch = 20, col = "blue"))
with(subset(res, logpval < -log10(0.05) & abs(logFC) > log2(threshold)), 
     points(logFC, logpval, pch = 20, col = "gray"))

abline(h = -log10(0.05), col = "red", lwd = 3)
abline(v = log2(1/threshold), col = "grey", lty = 2)
abline(v = log2(threshold), col = "grey", lty = 2)
UP <- nrow(subset(res, logFC > log2(threshold) & logpval > -log10(0.05)))
DO <- nrow(subset(res, logFC < log2(1/threshold) & logpval > -log10(0.05)))
legend("topleft", legend = c(parse(text = sprintf("Upregulated : %s",UP)), 
                           parse(text = sprintf("Downregulated : %s",DO))),
       col = c("red", "blue"), pch = 16, cex = 0.8)
```

I chose the FC threshold of a 25% change is often large enough to reflect real biological shifts, especially in metabolic pathways. This is a moderate and commonly accepted threshold in metabolomics, especially when we want to balance sensitivity and specificity. 
It captures more hits while still avoiding tiny, likely insignificant changes; which is good for pathway enrichment. 

#### Significant variables

According to the selection criteria mentioned above, **`r UP`**
variables are significantly up and **`r DO`** variables are
significantly down.

## Volcano plot for q values

```{r}
FC <- as.numeric(test[, "foldchange"])
Pvalue <- as.numeric(test[, "pvalue adjusted"])
threshold = 1.25
logFC <- log2(FC)
wh <- which(Pvalue == 0)
Pvalue[wh] <- 0.000001
logpval <- -log10(Pvalue)
res <- data.frame(cbind(logFC, logpval))
  
with(res, plot(logFC, logpval, pch = 20, ylab = "p_value (in -log10)", 
               xlab = "Fold change (in log2)"),cex.axis = 3,cex.lab = 3)
with(subset(res, logpval < -log10(0.05) ), 
     points(logFC, logpval, pch = 20, col = "black"))
with(subset(res, logFC > log2(1) & logpval > -log10(0.05)), 
     points(logFC,  logpval, pch = 20, col = "lightcoral"))
with(subset(res, logFC < log2(1) & logpval > -log10(0.05)), 
     points(logFC,  logpval, pch = 20, col = "lightblue"))
with(subset(res, logFC > log2(threshold) & logpval > -log10(0.05)), 
     points(logFC,  logpval, pch = 20, col = "red"))
with(subset(res, logFC < log2(1/threshold) & logpval > -log10(0.05)), 
     points(logFC,  logpval, pch = 20, col = "blue"))
with(subset(res, logpval < -log10(0.05) & abs(logFC) > log2(threshold)), 
     points(logFC, logpval, pch = 20, col = "gray"))

abline(h = -log10(0.05), col = "red", lwd = 3)
abline(v = log2(1/threshold), col = "grey", lty = 2)
abline(v = log2(threshold), col = "grey", lty = 2)
UP <- nrow(subset(res, logFC > log2(threshold) & logpval > -log10(0.05)))
DO <- nrow(subset(res, logFC < log2(1/threshold) & logpval > -log10(0.05)))
legend("topleft", legend = c(parse(text = sprintf("Upregulated : %s",UP)), 
                           parse(text = sprintf("Downregulated : %s",DO))),
       col = c("red", "blue"), pch = 16, cex = 0.8)
```


```{r}
plot_volcano <- function(data, foldchange_col = "foldchange", 
                         pvalue_col = "pvalue", 
                         threshold_fc = 1.5, threshold_pval = 0.05, 
                         y_label = '-log10(P-value)',
                         annotation_label = "A)", title = "AD vs CN") {
  
  FC <- as.numeric(data[, foldchange_col])
  Pvalue <- as.numeric(data[, pvalue_col])
  Pvalue[Pvalue == 0] <- 0.000001
  logFC <- log2(FC)
  logpval <- -log10(Pvalue)
  res <- data.frame(
    logFC = logFC,
    logpval = logpval,
    Biomarker = data[, "variable"]
  )
  res$Color <- "grey"
  res$Color[res$logFC > log2(threshold_fc) & res$logpval > -log10(threshold_pval)] <- "red"
  res$Color[res$logFC < -log2(threshold_fc) & res$logpval > -log10(threshold_pval)] <- "blue"
  res$Color[res$logpval <= -log10(threshold_pval)] <- "black"

  res$label <- ifelse(res$logFC > log2(threshold_fc) & res$logpval > -log10(threshold_pval) | 
                        res$logFC < -log2(threshold_fc) & res$logpval > -log10(threshold_pval), 
                      res$Biomarker, NA)
  
  ggplot(res, aes(x = logFC, y = logpval, label = label, color = Color)) +
    geom_point(size = 3, alpha = 0.6) +
    scale_color_manual(values = c("grey" = "grey", "red" = "red", "blue" = "blue", "black" = "black")) +
    ggrepel::geom_text_repel(aes(color = Color), size = 4.5, box.padding = 0.6, 
                     alpha = 10) +
    labs(x = 'Fold Change (log2)', y = y_label, title = title) +
    cowplot::theme_cowplot() +
    theme(legend.position = 'none', plot.title = element_text(hjust = 0.5, size = 18, 
                                                              vjust = 1.5)) +
    geom_hline(yintercept = -log10(threshold_pval), linetype = "dashed", color = "gray") +
    geom_vline(xintercept = log2(threshold_fc), linetype = "dashed", color = "gray") +
    geom_vline(xintercept = -log2(threshold_fc), linetype = "dashed", color = "gray") +
    ylim(c(0, 3)) + 
    annotate("text", y = 3, x = -1.7, label = annotation_label, size = 7, 
             fontface = "bold", angle = 0, color = "black") -> pv_volcano
  
  return(pv_volcano)
}

test1 <- test
plot_volcano(test1, foldchange_col = "foldchange", pvalue_col = "pvalue",
             threshold_fc = 1.25, threshold_pval = 0.05, annotation_label = "A)",
             y_label = '-log10(P-value)',
             title = "AD vs CN") -> pv_volcano

plot_volcano(test1, foldchange_col = "foldchange", pvalue_col = "pvalue adjusted",
             threshold_fc = 1.25, threshold_pval = 0.05, annotation_label = "B)",
             y_label = '-log10(P-value adjusted)',
             title = "AD vs CN") -> qv_volcano

ggsave("pictures/volcano_plots.png", 
       gridExtra::arrangeGrob(pv_volcano, qv_volcano, ncol = 2), 
       width = 16, height = 8, dpi = 300)
```

#### Significant variables

According to the selection criteria mentioned above, **`r UP`**
variables are significantly up and **`r DO`** variables are
significantly down.

## AUC and Fold changes for each Biomarker

```{r aucfoldchangetable3}
CN <- data0_1[which(data0_1$Allgr == "CN"), ]
AD = data0_1[which(data0_1$Allgr == "AD"), ]
n = ncol(CN)
dat = rbind(CN, AD)
dataX = as.matrix(dat[, 8:n])
dataY = as.factor(dat$Allgr)
n = ncol(dataX)
AUC <- ADNIMetabo::AUC_indiv(dataX, dataY, n, "AD vs CN")
AUC <- as.data.frame(AUC)
AUC$auc <- as.numeric(AUC$auc)
n <- length(AUC$auc)
for (i in seq_len(nrow(AUC))) {
  if (AUC[i, 2] < 0.50) {
    AUC[i, 2] <- 1 - AUC[i, 2]
  }
}
test <- as.data.frame(test)
test <- test |> 
    dplyr::rename(biomarker = variable, 
         CN_Mean = "mean group2", CN_SD = "sd group2", 
         AD_Mean = "mean group1", AD_SD = "sd group1", 
         pvalues = pvalue, foldchange = foldchange, 
         qvalues = "pvalue adjusted") |>
    dplyr::select_if(~sum(!is.na(.)) > 0)
final <- merge(AUC, test, by = "biomarker", all = FALSE )
final <- data.frame(final)
final$auc = as.numeric(final$auc)
final$AD_Mean = as.numeric(final$AD_Mean)
final$AD_SD = as.numeric(final$AD_SD)
final$CN_Mean = as.numeric(final$CN_Mean)
final$CN_SD = as.numeric(final$CN_SD)
final$pvalues = as.numeric(final$pvalues)
final$foldchange = as.numeric(final$foldchange)
final$qvalues = as.numeric(final$qvalues)
final[, -1] <- round(final[, -1] |> sapply(as.numeric), 3)
DT::datatable(final)
write.csv(final, "finaltableADvsCN.csv", row.names = F)
```

## List of significant biomarker after gender and Age effects adjusted

```{r}
b1 = as.data.frame(final[which(final$auc > 0.60), 1])
b1 = as.character(b1[, 1])
b1
newdata = data0_1[, c(1:7)]
c = colnames(data0_1)
for (j in 1:length(b1)) {
  newdata = cbind(newdata, data0_1[, which(c == b1[j])])
}
colnames(newdata)[-c(1:7)] = b1
data2 = as.data.frame(newdata)
significantadjbiomarkers <- data2[, -c(1:7)]
```

## Correlations of Significant Biomarkers

```{r, fig.dim = c(8, 8)}
if (ncol(significantadjbiomarkers) > 0) {
  correlations <- cor(as.matrix(significantadjbiomarkers), 
                      use = "pairwise.complete.obs")
  corrplot::corrplot(
    correlations, method = "square", order = "hclust", number.font = 5,
    main = "Correlations between significant biomarkers ",
    tl.cex = 0.5, mar = c(0, 0, 3, 0), tl.col = 1
  )
}
```


## Boxplot for significant Biomarkers for Age, gender adjusted data

```{r, fig.dim = c(8, 8)}
if (ncol(significantadjbiomarkers) > 0) {
  data2 <- as.data.frame(data2)
  sigbiomarkersadj <- colnames(data2)[8:length(data2)]
  metadata <- c("Gender", "Allgr", "APOE4", "Age")
  
  long_datasetsig <- melt(
    data2, id.vars = c('RID' , metadata), variable.name = "ASSAY", 
    value.name = "value", measure.vars = sigbiomarkersadj
  )
  
  long_datasetsig <- subset(long_datasetsig, Cohort != "HC")
  
  ggplot(long_datasetsig, aes(x = Cohort, col = Cohort, y = value)) +
    theme_minimal() +
    geom_boxplot() + facet_wrap(ASSAY ~ ., scales = "free") + 
    scale_color_discrete(guide = F) + 
    ggtitle("Biomarker values with significant group effect") +
    theme(strip.text = element_text(size = rel(0.6)))
}
```
