---
title: "ADNI Baseline Analysis: Alzheimer's Disease vs Cognitively Normal Case-Control Study"
author: "Analysis Report"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    fig_width: 8
    fig_height: 5
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, 
                     fig.align = "center", dpi = 300)

# Load required libraries
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(corrplot)
library(tableone)
library(RColorBrewer)
library(viridis)
library(plotly)
library(DT)
library(kableExtra)
library(ggpubr)
library(broom)
library(car)
library(effsize)
devtools::load_all()

# Set theme for plots
theme_set(theme_bw() + 
          theme(plot.title = element_text(hjust = 0.5, face = "bold"),
                legend.position = "bottom"))

# Custom color palette
custom_colors <- c("#2E86AB", "#A23B72", "#F18F01", "#C73E1D")
```

# Executive Summary

This analysis examines baseline differences between Alzheimer's Disease (AD) patients and Cognitively Normal (CN) controls in the ADNI dataset. The study includes demographic, clinical, neuropsychological, and biomarker comparisons.

# Data Preparation

```{r data-prep}
# Assuming your data preparation code from the original script
# Create baseline subset for case-control analysis
baseline_data <- adnim %>%
  filter(VISCODE == "bl", DX_bl %in% c("AD", "CN")) %>%
  mutate(
    Group = factor(DX_bl, levels = c("CN", "AD")),
    APOE4_status = case_when(
      APOE4 == 0 ~ "No APOE4",
      APOE4 == 1 ~ "One APOE4",
      APOE4 == 2 ~ "Two APOE4",
      TRUE ~ "Unknown"
    ),
    Education_level = case_when(
      PTEDUCAT < 12 ~ "< High School",
      PTEDUCAT == 12 ~ "High School",
      PTEDUCAT > 12 & PTEDUCAT < 16 ~ "Some College",
      PTEDUCAT >= 16 ~ "College+",
      TRUE ~ "Unknown"
    ),
    # Convert biomarker strings to numeric
    ABETA_numeric = as.numeric(ifelse(ABETA_bl == "", NA, ABETA_bl)),
    TAU_numeric = as.numeric(ifelse(TAU_bl == "", NA, TAU_bl))
  ) %>%
  select(-ABETA_bl, -TAU_bl, -ABETA, -TAU) # Remove string versions

# Summary of sample sizes
cat("Sample Sizes:\n")
table(baseline_data$Group)
```

# Demographic Characteristics

## Table 1: Baseline Demographics and Clinical Characteristics

```{r demographics-table}
# Define variables for Table 1
demo_vars <- c("AGE", "PTGENDER", "PTEDUCAT", "Education_level", 
               "APOE4", "APOE4_status")

# Create Table 1
tab1 <- CreateTableOne(vars = demo_vars, 
                       strata = "Group", 
                       data = baseline_data,
                       test = TRUE)

# Print formatted table
kable(print(tab1, showAllLevels = TRUE, printToggle = FALSE), 
      caption = "Table 1: Baseline Demographics") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Demographic Visualizations

```{r demo-plots, fig.height=8, fig.width=12}
# Age distribution
p1 <- ggplot(baseline_data, aes(x = Group, y = AGE, fill = Group)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3) +
  scale_fill_manual(values = custom_colors[1:2]) +
  labs(title = "Age Distribution by Group", y = "Age (years)") +
  theme(legend.position = "none")

# Gender distribution
p2 <- baseline_data %>%
  count(Group, PTGENDER) %>%
  group_by(Group) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = Group, y = prop, fill = PTGENDER)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = custom_colors[3:4]) +
  labs(title = "Gender Distribution by Group", y = "Proportion", fill = "Gender")

# Education distribution
p3 <- baseline_data %>%
  count(Group, Education_level) %>%
  ggplot(aes(x = Education_level, y = n, fill = Group)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = custom_colors[1:2]) +
  labs(title = "Education Distribution by Group", 
       x = "Education Level", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# APOE4 distribution
p4 <- baseline_data %>%
  count(Group, APOE4_status) %>%
  group_by(Group) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = APOE4_status, y = prop, fill = Group)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = custom_colors[1:2]) +
  labs(title = "APOE4 Status Distribution by Group", 
       x = "APOE4 Status", y = "Proportion") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

grid.arrange(p1, p2, p3, p4, ncol = 2)
```

# Neuropsychological Assessment Comparisons

## Table 2: Cognitive Test Scores

```{r cognitive-table}
# Define cognitive variables
cog_vars <- c("MMSE_bl", "ADAS11_bl", "ADAS13_bl", "ADASQ4_bl", 
              "LDELTOTAL_BL", "RAVLT_immediate_bl", "RAVLT_learning_bl", 
              "RAVLT_forgetting_bl")

# Create cognitive assessment table
tab2 <- CreateTableOne(vars = cog_vars, 
                       strata = "Group", 
                       data = baseline_data,
                       test = TRUE)

kable(print(tab2, printToggle = FALSE), 
      caption = "Table 2: Neuropsychological Assessment Scores") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Cognitive Assessment Visualizations

```{r cognitive-plots, fig.height=10, fig.width=14}
# Prepare data for plotting
cog_data <- baseline_data %>%
  select(Group, all_of(cog_vars)) %>%
  gather(key = "Test", value = "Score", -Group) %>%
  filter(!is.na(Score)) %>%
  mutate(
    Test_Label = case_when(
      Test == "MMSE_bl" ~ "MMSE",
      Test == "ADAS11_bl" ~ "ADAS-11",
      Test == "ADAS13_bl" ~ "ADAS-13", 
      Test == "ADASQ4_bl" ~ "ADAS-Q4",
      Test == "LDELTOTAL_BL" ~ "Logical Memory",
      Test == "RAVLT_immediate_bl" ~ "RAVLT Immediate",
      Test == "RAVLT_learning_bl" ~ "RAVLT Learning",
      Test == "RAVLT_forgetting_bl" ~ "RAVLT Forgetting",
      TRUE ~ Test
    )
  )

# Box plots for cognitive tests
ggplot(cog_data, aes(x = Group, y = Score, fill = Group)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 0.5) +
  scale_fill_manual(values = custom_colors[1:2]) +
  facet_wrap(~ Test_Label, scales = "free_y", ncol = 4) +
  labs(title = "Cognitive Test Scores by Group", 
       y = "Score") +
  theme(legend.position = "none", 
        strip.text = element_text(face = "bold"))
```

## Effect Sizes for Cognitive Tests

```{r effect-sizes}
# Calculate effect sizes (Cohen's d)
effect_sizes <- cog_data %>%
  group_by(Test_Label) %>%
  summarise(
    cohens_d = cohen.d(Score ~ Group, na.rm = TRUE)$estimate,
    ci_lower = cohen.d(Score ~ Group, na.rm = TRUE)$conf.int[1],
    ci_upper = cohen.d(Score ~ Group, na.rm = TRUE)$conf.int[2],
    p_value = t.test(Score ~ Group, na.rm = TRUE)$p.value,
    .groups = 'drop'
  ) %>%
  arrange(desc(abs(cohens_d)))

# Plot effect sizes
ggplot(effect_sizes, aes(x = reorder(Test_Label, abs(cohens_d)), y = cohens_d)) +
  geom_col(aes(fill = abs(cohens_d) > 0.8), alpha = 0.8) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
  coord_flip() +
  scale_fill_manual(values = c("FALSE" = custom_colors[1], "TRUE" = custom_colors[2]),
                   name = "Large Effect\n(|d| > 0.8)") +
  labs(title = "Effect Sizes (Cohen's d) for Cognitive Tests", 
       subtitle = "AD vs CN Comparison",
       x = "Cognitive Test", y = "Cohen's d (95% CI)") +
  geom_hline(yintercept = c(-0.8, -0.5, -0.2, 0.2, 0.5, 0.8), 
             linetype = "dashed", alpha = 0.5)

# Display effect sizes table
kable(effect_sizes, digits = 3, 
      caption = "Effect Sizes for Cognitive Test Differences") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# Biomarker Analysis

## Table 3: Biomarker Comparisons

```{r biomarker-table}
# Define biomarker variables
bio_vars <- c("FDG_bl", "AV45_bl", "PIB_bl", "ABETA_numeric", "TAU_numeric", "ICV_bl")

# Create biomarker table
tab3 <- CreateTableOne(vars = bio_vars, 
                       strata = "Group", 
                       data = baseline_data,
                       test = TRUE)

kable(print(tab3, printToggle = FALSE), 
      caption = "Table 3: Biomarker Comparisons") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Biomarker Visualizations

```{r biomarker-plots, fig.height=8, fig.width=12}
bio_data <- baseline_data %>%
  select(Group, FDG_bl, AV45_bl, #PIB_bl, 
         ABETA_numeric, TAU_numeric) %>%
  gather(key = "Biomarker", value = "Value", -Group) %>%
  filter(!is.na(Value)) %>%
  mutate(
    Biomarker_Label = case_when(
      Biomarker == "FDG_bl" ~ "FDG-PET",
      Biomarker == "AV45_bl" ~ "Amyloid PET (AV45)",
      # Biomarker == "PIB_bl" ~ "Amyloid PET (PIB)",
      Biomarker == "ABETA_numeric" ~ "CSF Aβ42",
      Biomarker == "TAU_numeric" ~ "CSF Tau",
      TRUE ~ Biomarker
    )
  )

p1 <- ggplot(bio_data, aes(x = Group, y = Value, fill = Group)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 0.5) +
  scale_fill_manual(values = custom_colors[1:2]) +
  facet_wrap(~ Biomarker_Label, scales = "free_y", ncol = 2) +
  labs(title = "Biomarker Levels by Group", y = "Value") +
  theme(legend.position = "none")

print(p1)

# corr. between biomarkers
bio_corr_data <- baseline_data %>%
  select(FDG_bl, AV45_bl, #PIB_bl, 
         ABETA_numeric, TAU_numeric) %>%
  rename("FDG-PET" = FDG_bl, 
         "Amyloid PET (AV45)" = AV45_bl,
         # "Amyloid PET (PIB)" = PIB_bl,
         "CSF Aβ42" = ABETA_numeric,
         "CSF Tau" = TAU_numeric)

corrplot(cor(bio_corr_data, use = "complete.obs"), 
         method = "color", type = "upper", 
         title = "Biomarker Correlations",
         mar = c(0,0,2,0))
```

# Neuroimaging Analysis: Structural MRI

## Overview of Neuroimaging Data

The ADNI dataset includes comprehensive structural MRI measurements derived from FreeSurfer analysis, providing detailed information about brain regional volumes and cortical thickness measures.

```{r imaging-data-prep}
# Add imaging variables to baseline data
baseline_data <- baseline_data %>%
  mutate(
    # Normalize brain volumes by ICV to account for head size differences
    Ventricles_norm = Ventricles_bl / ICV_bl * 1000,
    Hippocampus_norm = Hippocampus_bl / ICV_bl * 1000,
    WholeBrain_norm = WholeBrain_bl / ICV_bl * 1000,
    Entorhinal_norm = Entorhinal_bl / ICV_bl * 1000,
    Fusiform_norm = Fusiform_bl / ICV_bl * 1000,
    MidTemp_norm = MidTemp_bl / ICV_bl * 1000,
    
    # Create atrophy indices (higher values = more atrophy)
    Hippocampal_Atrophy = 1 - (Hippocampus_norm / max(Hippocampus_norm, na.rm = TRUE)),
    Entorhinal_Atrophy = 1 - (Entorhinal_norm / max(Entorhinal_norm, na.rm = TRUE)),
    
    # Ventricle-to-brain ratio (VBR)
    VBR = Ventricles_bl / WholeBrain_bl
  )

# Summary of imaging data availability
imaging_summary <- baseline_data %>%
  summarise(
    across(c(Ventricles_bl, Hippocampus_bl, WholeBrain_bl, 
             Entorhinal_bl, Fusiform_bl, MidTemp_bl, ICV_bl), 
           ~sum(!is.na(.)), .names = "n_{.col}"),
    .by = Group
  )

kable(imaging_summary, caption = "Structural MRI Data Availability") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Table 4: Structural MRI Comparisons

```{r imaging-table}
# Define imaging variables
imaging_vars <- c("ICV_bl", "Ventricles_norm", "Hippocampus_norm", "WholeBrain_norm", 
                  "Entorhinal_norm", "Fusiform_norm", "MidTemp_norm", "VBR")

# Create imaging table
tab4 <- CreateTableOne(vars = imaging_vars, 
                       strata = "Group", 
                       data = baseline_data,
                       test = TRUE)

kable(print(tab4, printToggle = FALSE), 
      caption = "Table 4: Structural MRI Measurements (ICV-normalized)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Regional Brain Volume Comparisons

```{r regional-volumes, fig.height=10, fig.width=14}
# Prepare data for regional volume analysis
regional_data <- baseline_data %>%
  select(Group, Ventricles_norm, Hippocampus_norm, WholeBrain_norm, 
         Entorhinal_norm, Fusiform_norm, MidTemp_norm) %>%
  gather(key = "Region", value = "Volume", -Group) %>%
  filter(!is.na(Volume)) %>%
  mutate(
    Region_Label = case_when(
      Region == "Ventricles_norm" ~ "Ventricles",
      Region == "Hippocampus_norm" ~ "Hippocampus",
      Region == "WholeBrain_norm" ~ "Whole Brain",
      Region == "Entorhinal_norm" ~ "Entorhinal Cortex",
      Region == "Fusiform_norm" ~ "Fusiform Gyrus",
      Region == "MidTemp_norm" ~ "Middle Temporal",
      TRUE ~ Region
    ),
    Region_Category = case_when(
      Region %in% c("Hippocampus_norm", "Entorhinal_norm") ~ "Memory-Related",
      Region %in% c("Fusiform_norm", "MidTemp_norm") ~ "Temporal Cortex",
      Region == "WholeBrain_norm" ~ "Global Measure",
      Region == "Ventricles_norm" ~ "CSF Space",
      TRUE ~ "Other"
    )
  )

# Regional volume box plots
p1 <- ggplot(regional_data, aes(x = Group, y = Volume, fill = Group)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  geom_jitter(width = 0.2, alpha = 0.2, size = 0.3) +
  scale_fill_manual(values = custom_colors[1:2]) +
  facet_wrap(~ Region_Label, scales = "free_y", ncol = 3) +
  labs(title = "Regional Brain Volumes by Diagnostic Group", 
       subtitle = "Volumes normalized by intracranial volume (ICV)",
       y = "Normalized Volume (per 1000 ICV units)") +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold"))

print(p1)

# Effect sizes for brain regions
regional_effects <- regional_data %>%
  group_by(Region_Label, Region_Category) %>%
  summarise(
    cohens_d = cohen.d(Volume ~ Group, na.rm = TRUE)$estimate,
    ci_lower = cohen.d(Volume ~ Group, na.rm = TRUE)$conf.int[1],
    ci_upper = cohen.d(Volume ~ Group, na.rm = TRUE)$conf.int[2],
    p_value = t.test(Volume ~ Group, na.rm = TRUE)$p.value,
    .groups = 'drop'
  ) %>%
  arrange(desc(abs(cohens_d)))

# Effect sizes plot
p2 <- ggplot(regional_effects, aes(x = reorder(Region_Label, abs(cohens_d)), y = cohens_d)) +
  geom_col(aes(fill = Region_Category), alpha = 0.8) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
  coord_flip() +
  scale_fill_viridis_d(name = "Region Category") +
  labs(title = "Effect Sizes for Regional Brain Volume Differences", 
       subtitle = "AD vs CN Comparison (Cohen's d with 95% CI)",
       x = "Brain Region", y = "Cohen's d") +
  geom_hline(yintercept = c(-1.2, -0.8, -0.5, -0.2, 0, 0.2, 0.5, 0.8, 1.2), 
             linetype = "dashed", alpha = 0.3)

print(p2)

# Display effect sizes table
kable(regional_effects, digits = 3, 
      caption = "Effect Sizes for Regional Brain Volume Differences") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Advanced Neuroimaging Analysis

### Brain Volume Ratios and Composite Scores

```{r brain-ratios, fig.height=8, fig.width=12}
# Calculate meaningful brain ratios
baseline_data <- baseline_data %>%
  mutate(
    # Hippocampus to total brain ratio
    Hip_to_Brain = Hippocampus_bl / WholeBrain_bl,
    
    # Entorhinal to hippocampus ratio
    Ento_to_Hip = Entorhinal_bl / Hippocampus_bl,
    
    # Temporal lobe composite (fusiform + middle temporal + entorhinal + hippocampus)
    Temporal_Composite = (Fusiform_norm + MidTemp_norm + Entorhinal_norm + Hippocampus_norm),
    
    # Atrophy severity index combining key AD regions
    AD_Atrophy_Index = scale(Hippocampal_Atrophy)[,1] + 
                      scale(Entorhinal_Atrophy)[,1] + 
                      scale(VBR)[,1]
  )

# Visualize ratios
ratio_data <- baseline_data %>%
  select(Group, Hip_to_Brain, Ento_to_Hip, VBR, AD_Atrophy_Index) %>%
  gather(key = "Ratio", value = "Value", -Group) %>%
  filter(!is.na(Value)) %>%
  mutate(
    Ratio_Label = case_when(
      Ratio == "Hip_to_Brain" ~ "Hippocampus/Brain Ratio",
      Ratio == "Ento_to_Hip" ~ "Entorhinal/Hippocampus Ratio", 
      Ratio == "VBR" ~ "Ventricle/Brain Ratio",
      Ratio == "AD_Atrophy_Index" ~ "AD Atrophy Index",
      TRUE ~ Ratio
    )
  )

ggplot(ratio_data, aes(x = Group, y = Value, fill = Group)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 0.5) +
  scale_fill_manual(values = custom_colors[1:2]) +
  facet_wrap(~ Ratio_Label, scales = "free_y", ncol = 2) +
  labs(title = "Brain Volume Ratios and Composite Measures", 
       y = "Ratio/Index Value") +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold"))

# Statistical tests for ratios
ratio_tests <- ratio_data %>%
  group_by(Ratio_Label) %>%
  summarise(
    AD_mean = mean(Value[Group == "AD"], na.rm = TRUE),
    CN_mean = mean(Value[Group == "CN"], na.rm = TRUE),
    p_value = t.test(Value ~ Group, na.rm = TRUE)$p.value,
    cohens_d = cohen.d(Value ~ Group, na.rm = TRUE)$estimate,
    .groups = 'drop'
  )

kable(ratio_tests, digits = 4, 
      caption = "Statistical Comparison of Brain Volume Ratios") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Correlation Analysis: Brain Structure and Cognition

```{r brain-cognition-correlation, fig.height=10, fig.width=14}
# Correlation between brain measures and cognitive scores
brain_cog_data <- baseline_data %>%
  select(Group, MMSE_bl, ADAS11_bl, ADAS13_bl, LDELTOTAL_BL, RAVLT_immediate_bl,
         Hippocampus_norm, Entorhinal_norm, Fusiform_norm, MidTemp_norm, 
         VBR, AD_Atrophy_Index) %>%
  na.omit()

# Correlation matrix
cor_matrix <- cor(brain_cog_data %>% select(-Group), use = "complete.obs")

# Enhanced correlation plot
corrplot(cor_matrix, method = "color", type = "upper", 
         order = "hclust", tl.cex = 0.8, tl.col = "black",
         title = "Correlations: Brain Structure and Cognitive Measures",
         mar = c(0,0,2,0))

# Scatter plots for key relationships
key_relationships <- data.frame(
  MMSE = brain_cog_data$MMSE_bl,
  ADAS11 = brain_cog_data$ADAS11_bl,
  Hippocampus = brain_cog_data$Hippocampus_norm,
  Entorhinal = brain_cog_data$Entorhinal_norm,
  VBR = brain_cog_data$VBR,
  Group = brain_cog_data$Group
)

# MMSE vs Hippocampus
p1 <- ggplot(key_relationships, aes(x = Hippocampus, y = MMSE, color = Group)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_color_manual(values = custom_colors[1:2]) +
  labs(title = "MMSE Score vs Hippocampal Volume",
       x = "Hippocampal Volume (ICV-normalized)",
       y = "MMSE Score") +
  theme(legend.position = "bottom")

# ADAS11 vs Entorhinal
p2 <- ggplot(key_relationships, aes(x = Entorhinal, y = ADAS11, color = Group)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_color_manual(values = custom_colors[1:2]) +
  labs(title = "ADAS-11 Score vs Entorhinal Volume",
       x = "Entorhinal Volume (ICV-normalized)", 
       y = "ADAS-11 Score") +
  theme(legend.position = "bottom")

# MMSE vs VBR
p3 <- ggplot(key_relationships, aes(x = VBR, y = MMSE, color = Group)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_color_manual(values = custom_colors[1:2]) +
  labs(title = "MMSE Score vs Ventricle-to-Brain Ratio",
       x = "Ventricle-to-Brain Ratio",
       y = "MMSE Score") +
  theme(legend.position = "bottom")

# Combined atrophy index vs cognition
p4 <- ggplot(brain_cog_data, aes(x = AD_Atrophy_Index, y = MMSE_bl, color = Group)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_color_manual(values = custom_colors[1:2]) +
  labs(title = "MMSE Score vs AD Atrophy Index",
       x = "AD Atrophy Index (standardized)",
       y = "MMSE Score") +
  theme(legend.position = "bottom")

grid.arrange(p1, p2, p3, p4, ncol = 2)
```

### Machine Learning Classification using Neuroimaging

```{r ml-imaging, fig.height=8, fig.width=12}
# Prepare data for machine learning
ml_data <- baseline_data %>%
  select(Group, AGE, Male, PTEDUCAT, APOE4_1, APOE4_2,
         Hippocampus_norm, Entorhinal_norm, Fusiform_norm, MidTemp_norm,
         Ventricles_norm, VBR, AD_Atrophy_Index) %>%
  mutate(AD = ifelse(Group == "AD", 1, 0)) %>%
  na.omit()

# Imaging-only model
imaging_model <- glm(AD ~ Hippocampus_norm + Entorhinal_norm + Fusiform_norm + 
                     MidTemp_norm + Ventricles_norm + VBR,
                     data = ml_data, family = "binomial")

# Combined model (demographics + imaging)
combined_model <- glm(AD ~ AGE + Male + PTEDUCAT + APOE4_1 + APOE4_2 +
                      Hippocampus_norm + Entorhinal_norm + Fusiform_norm + 
                      MidTemp_norm + Ventricles_norm + VBR,
                      data = ml_data, family = "binomial")

# Model comparison
imaging_results <- tidy(imaging_model) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    OR = exp(estimate),
    CI_lower = exp(estimate - 1.96 * std.error),
    CI_upper = exp(estimate + 1.96 * std.error),
    Model = "Imaging Only"
  )

combined_results <- tidy(combined_model) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    OR = exp(estimate),
    CI_lower = exp(estimate - 1.96 * std.error),
    CI_upper = exp(estimate + 1.96 * std.error),
    Model = "Combined"
  )

# Model performance comparison
imaging_auc <- pROC::auc(pROC::roc(ml_data$AD, predict(imaging_model, type = "response")))
combined_auc <- pROC::auc(pROC::roc(ml_data$AD, predict(combined_model, type = "response")))

cat("Model Performance (AUC):\n")
cat("Imaging Only:", round(imaging_auc, 3), "\n")
cat("Combined Model:", round(combined_auc, 3), "\n")

# ROC Curves
roc_imaging <- pROC::roc(ml_data$AD, predict(imaging_model, type = "response"))
roc_combined <- pROC::roc(ml_data$AD, predict(combined_model, type = "response"))

roc_data <- data.frame(
  Sensitivity = c(roc_imaging$sensitivities, roc_combined$sensitivities),
  Specificity = c(1 - roc_imaging$specificities, 1 - roc_combined$specificities),
  Model = c(rep("Imaging Only", length(roc_imaging$sensitivities)),
            rep("Combined", length(roc_combined$sensitivities)))
)

ggplot(roc_data, aes(x = Specificity, y = Sensitivity, color = Model)) +
  geom_line(size = 1.2) +
  geom_abline(linetype = "dashed", alpha = 0.5) +
  scale_color_manual(values = custom_colors[1:2]) +
  labs(title = "ROC Curves: Imaging vs Combined Models",
       subtitle = paste0("Imaging AUC = ", round(imaging_auc, 3), 
                        ", Combined AUC = ", round(combined_auc, 3)),
       x = "1 - Specificity", y = "Sensitivity") +
  theme(legend.position = "bottom")
```

## Neuroimaging Summary and Clinical Relevance

```{r imaging-summary}
# Key imaging findings summary
cat("NEUROIMAGING KEY FINDINGS:\n")
cat("==========================\n\n")

# Hippocampal atrophy
hip_test <- t.test(Hippocampus_norm ~ Group, data = baseline_data)
cat("Hippocampal Atrophy:\n")
cat("  AD patients show", round((1 - hip_test$estimate[1]/hip_test$estimate[2])*100, 1), 
    "% volume reduction\n")
cat("  Effect size (Cohen's d):", 
    round(cohen.d(baseline_data$Hippocampus_norm ~ baseline_data$Group, na.rm = TRUE)$estimate, 2), "\n")
cat("  p-value:", format.pval(hip_test$p.value), "\n\n")

# Entorhinal atrophy  
ento_test <- t.test(Entorhinal_norm ~ Group, data = baseline_data)
cat("Entorhinal Cortex Atrophy:\n")
cat("  AD patients show", round((1 - ento_test$estimate[1]/ento_test$estimate[2])*100, 1), 
    "% volume reduction\n")
cat("  Effect size (Cohen's d):", 
    round(cohen.d(baseline_data$Entorhinal_norm ~ baseline_data$Group, na.rm = TRUE)$estimate, 2), "\n")
cat("  p-value:", format.pval(ento_test$p.value), "\n\n")

# Ventricular enlargement
vent_test <- t.test(Ventricles_norm ~ Group, data = baseline_data)
cat("Ventricular Enlargement:\n")
cat("  AD patients show", round((vent_test$estimate[1]/vent_test$estimate[2] - 1)*100, 1), 
    "% volume increase\n")
cat("  Effect size (Cohen's d):", 
    round(cohen.d(baseline_data$Ventricles_norm ~ baseline_data$Group, na.rm = TRUE)$estimate, 2), "\n")
cat("  p-value:", format.pval(vent_test$p.value), "\n\n")

cat("Clinical Implications:\n")
cat("- Hippocampal and entorhinal atrophy are hallmark features of AD\n")
cat("- Ventricular enlargement reflects global brain atrophy\n")
cat("- Imaging measures strongly correlate with cognitive performance\n")
cat("- Combined imaging-clinical models achieve excellent diagnostic accuracy\n")
```

# Statistical Modeling

## Logistic Regression: Predictors of AD Diagnosis

```{r logistic-regression}
# data for modeling (complete cases only)
model_data <- baseline_data %>%
  select(Group, AGE, Male, PTEDUCAT, APOE4_1, APOE4_2,
         MMSE_bl, ADAS11_bl, FDG_bl, ABETA_numeric, TAU_numeric,
         Hippocampus_norm, Entorhinal_norm, VBR) %>%
  mutate(AD = ifelse(Group == "AD", 1, 0)) %>%
  na.omit()

cat("Model data dimensions:", dim(model_data), "\n")

# Univariate models
uni_vars <- c("AGE", "Male", "PTEDUCAT", "APOE4_1", "APOE4_2", 
              "MMSE_bl", "ADAS11_bl", "FDG_bl", #"MOCA_bl",  "CDRSB_bl", 
              "ABETA_numeric", "TAU_numeric",
              "Hippocampus_norm", "Entorhinal_norm", "VBR")

uni_results <- map_dfr(uni_vars, ~{
  formula_str <- paste("AD ~", .x)
  model <- glm(as.formula(formula_str), data = model_data, family = "binomial")
  tidy(model) %>%
    filter(term == .x) %>%
    mutate(variable = .x)
})

# Multivariate model
multi_model <- glm(AD ~ AGE + Male + PTEDUCAT + APOE4_1 + APOE4_2 + 
                   MMSE_bl + ADAS11_bl + #MOCA_bl +  CDRSB_bl +
                   FDG_bl + ABETA_numeric + TAU_numeric,
                   data = model_data, family = "binomial")

multi_results <- tidy(multi_model) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    OR = exp(estimate),
    CI_lower = exp(estimate - 1.96 * std.error),
    CI_upper = exp(estimate + 1.96 * std.error)
  )

kable(multi_results %>% 
      select(term, OR, CI_lower, CI_upper, p.value) %>%
      mutate(across(where(is.numeric), round, 3)),
      caption = "Multivariate Logistic Regression Results",
      col.names = c("Variable", "OR", "95% CI Lower", "95% CI Upper", "p-value")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

cat("Model AUC:", pROC::auc(pROC::roc(model_data$AD, predict(multi_model, type = "response"))), "\n")
```
## Forest Plot of Odds Ratios

```{r forest-plot, fig.height=8, fig.width=10}
# Create forest plot
forest_data <- multi_results %>%
  mutate(
    Variable_Label = case_when(
      term == "AGE" ~ "Age (per year)",
      term == "Male" ~ "Male gender",
      term == "PTEDUCAT" ~ "Education (per year)",
      term == "APOE4_1" ~ "APOE4 (1 allele)",
      term == "APOE4_2" ~ "APOE4 (2 alleles)",
      term == "MMSE_bl" ~ "MMSE score",
      term == "ADAS11_bl" ~ "ADAS-11 score",
      term == "FDG_bl" ~ "FDG-PET",
      term == "ABETA_numeric" ~ "CSF Aβ42",
      term == "TAU_numeric" ~ "CSF Tau",
      term == "Hippocampus_norm" ~ "Hippocampus_norm",
      TRUE ~ term
    ),
    Significant = ifelse(p.value < 0.05, "Significant", "Non-significant")
  )

ggplot(forest_data, aes(x = reorder(Variable_Label, OR), y = OR)) +
  geom_point(aes(color = Significant), size = 3) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper, color = Significant), 
                width = 0.2) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  coord_flip() +
  scale_color_manual(values = c("Significant" = custom_colors[2], 
                               "Non-significant" = custom_colors[1])) +
  scale_y_log10() +
  labs(title = "Forest Plot: Odds Ratios for AD Diagnosis",
       subtitle = "Multivariate Logistic Regression Model",
       x = "Variable", y = "Odds Ratio (95% CI, log scale)") +
  theme(legend.title = element_blank())
```

# Advanced Visualizations

## Principal Component Analysis

```{r pca-analysis, fig.height=8, fig.width=12}
# PCA on cognitive and biomarker variables
pca_data <- baseline_data %>%
  select(Group, MMSE_bl, ADAS11_bl, ADAS13_bl, LDELTOTAL_BL,
         RAVLT_immediate_bl, FDG_bl, ABETA_numeric, TAU_numeric) %>%
  na.omit()

# Perform PCA
pca_result <- prcomp(pca_data[,-1], scale = TRUE, center = TRUE)

# PCA plot
pca_df <- data.frame(pca_result$x) %>%
  mutate(Group = pca_data$Group)

p1 <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Group)) +
  geom_point(alpha = 0.6, size = 2) +
  stat_ellipse(type = "norm", linetype = 2) +
  scale_color_manual(values = custom_colors[1:2]) +
  labs(title = "Principal Component Analysis",
       subtitle = "Cognitive and Biomarker Variables",
       x = paste0("PC1 (", round(summary(pca_result)$importance[2,1]*100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_result)$importance[2,2]*100, 1), "%)"))

# Scree plot
scree_data <- data.frame(
  PC = 1:length(pca_result$sdev),
  Variance = pca_result$sdev^2 / sum(pca_result$sdev^2)
)

p2 <- ggplot(scree_data, aes(x = PC, y = Variance)) +
  geom_col(fill = custom_colors[3], alpha = 0.7) +
  geom_line(group = 1) +
  geom_point() +
  labs(title = "Scree Plot", y = "Proportion of Variance", x = "Principal Component")

grid.arrange(p1, p2, ncol = 2)

# Variable loadings
loadings_df <- data.frame(pca_result$rotation[,1:2]) %>%
  rownames_to_column("Variable") %>%
  mutate(
    Variable_Label = case_when(
      Variable == "MMSE_bl" ~ "MMSE",
      Variable == "ADAS11_bl" ~ "ADAS-11",
      Variable == "ADAS13_bl" ~ "ADAS-13",
      Variable == "LDELTOTAL_BL" ~ "Logical Memory",
      Variable == "RAVLT_immediate_bl" ~ "RAVLT Immediate",
      Variable == "FDG_bl" ~ "FDG-PET",
      Variable == "ABETA_numeric" ~ "CSF Aβ42",
      Variable == "TAU_numeric" ~ "CSF Tau",
      TRUE ~ Variable
    )
  )

# Biplot
ggplot() +
  geom_point(data = pca_df, aes(x = PC1, y = PC2, color = Group), alpha = 0.6) +
  geom_segment(data = loadings_df, 
               aes(x = 0, y = 0, xend = PC1*5, yend = PC2*5),
               arrow = arrow(length = unit(0.3, "cm")), color = "red") +
  geom_text(data = loadings_df, 
            aes(x = PC1*5.5, y = PC2*5.5, label = Variable_Label),
            size = 3, color = "red") +
  scale_color_manual(values = custom_colors[1:2]) +
  labs(title = "PCA Biplot", 
       subtitle = "Variables (red) and Individuals (colored by group)")
```

# Summary and Conclusions

```{r summary-stats}
# Summary statistics
cat("Key Findings:\n")
cat("=============\n\n")

# Sample sizes
n_ad <- sum(baseline_data$Group == "AD", na.rm = TRUE)
n_cn <- sum(baseline_data$Group == "CN", na.rm = TRUE)
cat("Sample Sizes: AD =", n_ad, ", CN =", n_cn, "\n\n")

# Age comparison
age_test <- t.test(AGE ~ Group, data = baseline_data)
cat("Age: AD patients are significantly older than CN controls\n")
cat("  Mean difference:", round(age_test$estimate[1] - age_test$estimate[2], 1), "years\n")
cat("  p-value:", format.pval(age_test$p.value, digits = 3), "\n\n")

# MMSE comparison
mmse_test <- t.test(MMSE_bl ~ Group, data = baseline_data)
cat("MMSE: AD patients have significantly lower scores\n")
cat("  Mean difference:", round(mmse_test$estimate[2] - mmse_test$estimate[1], 1), "points\n")
cat("  p-value:", format.pval(mmse_test$p.value, digits = 3), "\n\n")

# APOE4 association
apoe_test <- chisq.test(table(baseline_data$Group, baseline_data$APOE4))
cat("APOE4: Significantly associated with AD diagnosis\n")
cat("  Chi-square p-value:", format.pval(apoe_test$p.value, digits = 3), "\n")
```

## Key Statistical Differences

The analysis reveals significant differences between AD patients and CN controls across multiple domains:

**Demographics:**
- AD patients are significantly older than CN controls
- Higher proportion of APOE4 carriers in AD group
- No significant gender differences

**Cognitive Function:**
- AD patients show severe impairment on all cognitive measures
- Largest effect sizes observed for ADAS scores and memory tests
- MMSE scores clearly differentiate the groups

**Biomarkers:**
- Lower CSF Aβ42 levels in AD patients
- Higher CSF Tau levels in AD patients  
- Reduced FDG-PET metabolism in AD group

**Clinical Implications:**
The multivariate model identifies key predictors of AD diagnosis, with cognitive measures and biomarkers showing the strongest associations. These findings support the use of combined cognitive and biomarker assessments for AD diagnosis and monitoring.

---

*Analysis completed on `r Sys.Date()`*
*Data source: ADNI (Alzheimer's Disease Neuroimaging Initiative)*