---
title: "Additional file 2: Supplementary Materials"
subtitle: "WGCNA"
author: "Mukesha et al."
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    self_contained: true
    code_folding: hide
  word_document:
    toc: true
  pdf_document:
    toc: true
    latex_engine: xelatex
bibliography: AD_vs_CN_references.bib
editor_options:
  markdown:
    wrap: 72
---

# WGCNA analysis

WGCNA pipeline for metabolomics
    - Assumes you have data frames in your environment:
    - data_matrix (the full dataset: rows = samples, columns include 
    metabolite columns and clinical cols)
    - top_metabolite_table (columns: Metabolite, Class, LipidFlag)  
        -- metabolites selected by LASSO

```{r set-up}
required <- c("WGCNA","tidyverse","impute","pheatmap","viridis","ggpubr",
              "igraph","ggraph")
to_install <- required[!(required %in% installed.packages()[,"Package"])]
if (length(to_install)) install.packages(to_install, 
                                        repos = "https://cloud.r-project.org")

library(WGCNA)
library(tidyverse)
library(impute)
library(pheatmap)
library(viridis)
library(ggpubr)
library(igraph)
library(ggraph)

library(readr)

options(stringsAsFactors = FALSE)
enableWGCNAThreads() 

set.seed(12345)
```


## Prepare expression matrix

Choose which metabolites to use:

```{r load-data}
use_lasso_selected <- TRUE

data_full <- read_csv("onlyselectedADvsCN.csv") |>
    as_tibble()
sapply(colnames(data_full), ADNIMetabo::transform_colnames, 
                         USE.NAMES = FALSE)

top_metabolite_table <- read.csv("results/top_metabolite_table.csv") |>
    as_tibble()

if (use_lasso_selected) {
  met_list <- unique(top_metabolite_table$Metabolite)
  missing_mets <- setdiff(met_list, colnames(data_full))
  if (length(missing_mets) > 0) {
    message(
        "Warning: some metabolites from top_metabolite_table not found in data_full: ",
        paste(missing_mets, collapse = ", "))
  }
  expr_df <- data_full %>% dplyr::select(all_of(intersect(met_list, colnames(data_full))))
} else {
  clinical_cols <- c("RID","APOE4","Gender","Age","Allgr","Plate","V1")
  expr_df <- data_full %>% dplyr::select(where(is.numeric), -any_of(clinical_cols))
}

cat("Expression matrix dims (samples x metabolites):", dim(expr_df), "\n")

# transpose to WGCNA format: rows = samples, columns = genes (metabolites)
dataExpr <- as.data.frame(expr_df)
rownames(dataExpr) <- NULL # keep default numeric rownames for now
```

## Missing data handling

WGCNA tolerates some missingness but imputation recommended for metabolomics

```{r missing-data-handling}
na_frac <- sum(is.na(dataExpr)) / prod(dim(dataExpr))
cat(sprintf("Overall missing fraction: %.4f\n", na_frac))

# for more than tiny fraction, do kNN imputation (impute::impute.knn)
if (na_frac > 0) {
  # impute.knn expects matrix with 
  # -> rows = genes/metabolites, and 
  # -> cols = samples 
  # Then => transpose
  knn_in <- t(as.matrix(dataExpr))
  knn_out <- impute::impute.knn(knn_in, k = 10)
  datExpr <- as.data.frame(t(knn_out$data))
} else {
  datExpr <- dataExpr
}
```

## Remove metabolites with (near) zero variance

```{r rm-metabo-with-no-variance}
varv <- apply(datExpr, 2, var, na.rm = TRUE)
keep_cols <- names(varv)[varv > 1e-6]
datExpr <- datExpr[, keep_cols, drop = FALSE]
cat("After removing zero-variance metabolites, ncol =", ncol(datExpr), "\n")
```

```{r optional}
sample_names <- data_full$RID
if (is.null(sample_names)) sample_names <- paste0("S", seq_len(nrow(datExpr)))
rownames(datExpr) <- sample_names
```

## Prepare traits

```{r prepare-traits}
traits <- data_full %>% select(RID, APOE4, Gender, Age, Allgr) %>% as.data.frame()
rownames(traits) <- sample_names  # assume row order matches -> very important!
# if the order mismatches, i must align by RID.

# let's convert Group & Gender to numeric factors for correlation:
traits$Group <- as.numeric(factor(traits$Allgr, levels = c("CN","AD"))) # CN=1,AD=2 -> convert to 0/1
traits$Group <- traits$Group - 1
traits$APOE4 <- as.numeric(traits$APOE4)  # assuming numeric 0/1/2
traits$GenderNum <- as.numeric(factor(traits$Gender)) - 1 # e.g. MALE=0 FEMALE=1
clinical_traits <- traits %>% select(Group, APOE4, GenderNum, Age)
```

## Check for goodSamples/goodGenes/metabolites

```{r check-good-data}
gsg <- goodSamplesGenes(datExpr, verbose = 3)
if (!gsg$allOK) {
  # rm offending genes/samples
  if (sum(!gsg$goodGenes) > 0) {
    cat("Removing bad genes:", paste(colnames(datExpr)[!gsg$goodGenes], 
                                     collapse = ", "), "\n")
    datExpr <- datExpr[, gsg$goodGenes]
  }
  if (sum(!gsg$goodSamples) > 0) {
    cat("Removing bad samples:", paste(rownames(datExpr)[!gsg$goodSamples], 
                                       collapse = ", "), "\n")
    datExpr <- datExpr[gsg$goodSamples, ]
    clinical_traits <- clinical_traits[gsg$goodSamples, , drop = FALSE]
  }
}
```

## Choose soft-threshold power

For metabolomics, signed network often makes sense. 
I will compute recommended powers.

```{r soft-threshold}
powers <- c(c(1:10), seq(12, 20, 2))
sft <- pickSoftThreshold(datExpr,
    powerVector = powers,
    verbose = 5,
    networkType = "signed"
)

png("WGCNA_SoftThreshold.png", width = 900, height = 600, res = 150)
par(mfrow = c(1, 2))
plot(sft$fitIndices[, 1], 
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     xlab = "Soft Threshold (power)", 
     ylab = "Scale Free Topology Model Fit, signed R^2",
     main = "Scale independence"
)
abline(h = 0.8, col = "red", lty = 2)
plot(sft$fitIndices[, 1], sft$fitIndices[, 5],
    xlab = "Soft Threshold (power)", 
    ylab = "Mean connectivity", 
    main = "Mean connectivity"
)
dev.off()
```

Choose power:

automatic pick: smallest power with R^2 > 0.8, otherwise use 6 or 8

```{r choose-power}
powerTable <- sft$fitIndices
selPower <- min(powerTable$Power[ powerTable$SFT.R.sq > 0.6 ], na.rm = TRUE)
if (is.infinite(selPower) || is.na(selPower)) selPower <- 6
cat("Selected soft-threshold power:", selPower, "\n")
```

## Build adjacency and TOM, cluster genes

Use signed network; TOMType = "signed"

```{r build-adjacency-and-TOM}
adjacency <- adjacency(datExpr, power = selPower, type = "signed")
TOM <- TOMsimilarity(adjacency, TOMType = "signed")
dissTOM <- 1 - TOM

geneTree <- hclust(as.dist(dissTOM), method = "average")

# module detection with dynamic tree cut
minModuleSize <- 20   # for metabolites, can be smaller (e.g., 10) depending on data
dynamicMods <- cutreeDynamic(dendro = geneTree, distM = dissTOM,
                             deepSplit = 2, pamRespectsDendro = TRUE,
                             minClusterSize = minModuleSize)
moduleColors <- labels2colors(dynamicMods)

# merge close modules based on eigengene correlation
MEList <- moduleEigengenes(datExpr, colors = moduleColors)
MEs <- MEList$eigengenes
# calculate dissimilarity of module eigengenes
MEDiss <- 1 - cor(MEs)
METree <- hclust(as.dist(MEDiss), method = "average")
# merge threshold
mergeCutHeight <- 0.25
merge <- mergeCloseModules(datExpr, moduleColors, cutHeight = mergeCutHeight, verbose = 3)
mergedColors <- merge$colors
mergedMEs <- merge$newMEs

moduleColors <- mergedColors
MEs <- mergedMEs
```

## Visualize dendrogram + module colors

```{r}
png("WGCNA_Dendrogram_modules.png", width = 1400, height = 700, res = 180)
plotDendroAndColors(geneTree, moduleColors, "Dynamic Tree Cut + Merged",
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05,
                    main = "Metabolite clustering dendrogram")
dev.off()

```

## Module-trait relationships

```{r}
moduleTraitCor <- cor(MEs, clinical_traits, use = "p")
moduleTraitPvalue <- corPvalueStudent(moduleTraitCor, 
                                      nSamples = nrow(datExpr))

textMatrix <- paste0(signif(moduleTraitCor, 2), 
                     "\n(", signif(moduleTraitPvalue,2), ")")
png("WGCNA_ModuleTraitHeatmap.png", width = 800, height = 900, res = 150)
par(mar = c(6, 8, 3, 3))
labeledHeatmap(Matrix = moduleTraitCor,
               xLabels = colnames(clinical_traits),
               yLabels = names(MEs),
               ySymbols = names(MEs),
               colorLabels = FALSE,
               colors = colorRampPalette(
                   rev(brewer.pal(n = 7, name = "RdYlBu"))
                   )(50),
               textMatrix = textMatrix,
               setStdMargins = FALSE,
               cex.text = 0.8,
               zlim = c(-1,1),
               main = "Module-trait relationships (correlation (p-value))")
dev.off()
```

## Eigengene plots for key modules

pick top modules significantly associated with Group (AD)

```{r}
modCorDF <- tibble(
    Module = rownames(moduleTraitCor),
    corGroup = moduleTraitCor[, "Group"],
    pGroup = moduleTraitPvalue[, "Group"]
) %>%
    arrange(pGroup)
topMods <- modCorDF$Module[1:min(4, nrow(modCorDF))] # top 4

# Prepare eigengene data for plotting
MEs_df <- MEs %>% as.data.frame()
MEs_df$Sample <- rownames(datExpr)
MEs_long <- pivot_longer(MEs_df, cols = starts_with("ME"), 
                         names_to = "Module", values_to = "Eigengene")
MEs_long <- left_join(MEs_long, clinical_traits %>% 
                          mutate(Sample = rownames(clinical_traits)), 
                      by = c("Sample" = "Sample"))
# Keep only top modules
MEs_long_sub <- MEs_long %>% filter(Module %in% topMods)

pEig <- ggplot(MEs_long_sub, aes(x = factor(Group), y = Eigengene, 
                                 fill = factor(Group))) +
    geom_boxplot(outlier.size = 0.6, linewidth = 0.25) +
    facet_wrap(~Module, scales = "free_y", ncol = 2) +
    scale_fill_manual(values = c("0" = "#2166AC", "1" = "#B2182B"), 
                      labels = c("CN", "AD")) +
    theme_classic(base_size = 12) +
    labs(x = "", y = "Module eigengene", 
         title = "Module eigengenes by Group (AD vs CN)") +
    theme(legend.position = "none")
ggsave("WGCNA_Eigengene_boxplots.png", pEig, width = 8, 
       height = 6, dpi = 300)
```


## Identify hub metabolites in each module

calculate intramodular connectivity (kME) as correlation to module eigengene

```{r}
kME <- as.data.frame(signedKME(datExpr, MEs))
# For each module, list top hubs
hubs_list <- lapply(names(MEs), function(me){
  mname <- sub("^ME", "", me)
  kcol <- kME[[me]]
  dfh <- tibble(Metabolite = colnames(datExpr), kME = kcol,
                Module = mname) %>%
    arrange(desc(kME))
  head(dfh, 10)
})
names(hubs_list) <- names(MEs)

# Save hub table for the top module associated with Group
topModuleName <- topMods[1]
topModuleShort <- sub("^ME", "", topModuleName)
hubs_top <- hubs_list[[topModuleName]]
write.csv(hubs_top, paste0("WGCNA_Hubs_topModule_", 
                           topModuleShort, ".csv"), 
          row.names = FALSE)
```


```{r}
# ---------- 11. Visualize network of top module (top hub metabolites) ----------
# pick a module to show network (top associated with Group)
moduleGenes <- colnames(datExpr)[moduleColors == topModuleShort]
cat("Number of genes in top module", topModuleShort, ":", length(moduleGenes), "\n")
# If module big, pick top 30 by kME
topN <- min(30, length(moduleGenes))
kme_sub <- kME[moduleGenes, paste0("ME", topModuleShort), drop=FALSE]
topGenes <- rownames(kme_sub)[order(-kme_sub[[1]])][1:topN]

# Create subnetwork TOM for topGenes
subTOM <- TOM[topGenes, topGenes]
gmat <- graph_from_adjacency_matrix(subTOM, mode="undirected", weighted=TRUE, diag=FALSE)
V(gmat)$label <- topGenes
# node size by kME
kme_vals <- kME[topGenes, paste0("ME", topModuleShort)]
V(gmat)$size <- scales::rescale(kme_vals, to = c(3,8))
# color nodes by lipidFlag if available
if("LipidFlag" %in% colnames(top_metabolite_table)) {
  lipflag <- top_metabolite_table %>% filter(Metabolite %in% topGenes) %>% select(Metabolite, LipidFlag)
  lipmap <- setNames(lipflag$LipidFlag, lipflag$Metabolite)
  V(gmat)$lip <- lipmap[V(gmat)$label]
  # create palette
  pal <- c("lipid"="#2b8cbe","non-lipid"="#f03b20")
  V(gmat)$color <- pal[ as.character(V(gmat)$lip) ]
  V(gmat)$color[is.na(V(gmat)$color)] <- "grey40"
} else {
  V(gmat)$color <- "grey40"
}

# plot with ggraph
png(paste0("WGCNA_network_topModule_", topModuleShort, ".png"), width=900, height=900, res=150)
ggraph(gmat, layout = "fr") +
  geom_edge_link(aes(width = weight), alpha = 0.4, colour = "grey60") +
  geom_node_point(aes(size = size, color = I(V(gmat)$color))) +
  geom_node_text(aes(label=label), repel=TRUE, size=3) +
  theme_void() +
  ggtitle(paste("Top-module network:", topModuleShort))
dev.off()

# ---------- 12. Export for Cytoscape ----------
# Export edge and node tables for Cytoscape (for the top module)
cyt <- exportNetworkToCytoscape(subTOM,
                               edgeFile = paste0("CytoscapeInput-edges-", topModuleShort, ".txt"),
                               nodeFile = paste0("CytoscapeInput-nodes-", topModuleShort, ".txt"),
                               weighted = TRUE,
                               threshold = 0.1,    # edge weight threshold for Cytoscape export
                               nodeNames = topGenes,
                               nodeAttr = kme_vals)
cat("Cytoscape files written for module", topModuleShort, "\n")

# ---------- 13. Save workspace objects of interest ----------
save(list = c("datExpr", "moduleColors", "MEs", "hubs_list"), file = "WGCNA_results_metabolomics.RData")

cat("WGCNA analysis complete. Results and figures saved to working directory.\n")

```

